---
title: "Stomatal Behaviour 2026 - 01 Data Cleaning Script"
author: "Lena Sachsenmaier"
output:
  html_document:
    df_print: paged
contact: lena.sachsenmaier.research@gmail.com.de
created: "2026-07-30"
updated: "`r format(Sys.time(), '%d %B, %Y')`"
---


```{r setup}
library(here)
library(tidyverse)

```



# 1. Create folder structure

```{r create the folder structure needed}

dir.create("data/raw", recursive = TRUE)
dir.create("data/processed", recursive = TRUE)

dir.create("output/models", recursive = TRUE)
dir.create("output/figures", recursive = TRUE)
dir.create("output/tables", recursive = TRUE)


```


# 2. Data download
## 2.1. Stomatal conductance (& associated climate station data)

Primary data are currently restricted during peer review. Will be made publicly available upon acceptance of the manuscript.

```{r download main data set + climate station data}

data <- here(
  "data", "raw",
  "01_data_stomatal_conductance.csv"
)

climate_data <- here(
  "data", "raw",
  "02_data_climate_station_ARBOfun.csv"
)

if (!file.exists(data)) {
  message(
    "The primary data are currently restricted during peer review. ",
    "If you are a reviewer of this manuscript, please use the data access link ",
    "provided by the editor to download the dataset and place it in ",
    "'data/raw/' before proceeding."
  )
}
if (!file.exists(climate_data)) {
  message(
    "The primary data are currently restricted during peer review. ",
    "If you are a reviewer of this manuscript, please use the data access link ",
    "provided by the editor to download the dataset and place it in ",
    "'data/raw/' before proceeding."
  )
}


```

## 2.2. Leaf carbon isotope data

Kahl, A., Schnabel, F., Richter, R., Olbrich, J., Zahner, G., Weigelt, A., & Wirth, C. (2026). 
Leaf δ¹³C of 38 tree species measured in 2018 and 2021 (ARBOfun research arboretum, Germany) 
[Dataset]. Zenodo. https://doi.org/10.5281/zenodo.22284412

Primary data are currently restricted during peer review. Will be made publicly available upon acceptance of the manuscript.

```{r download d13C data}

delta_d13C_species_means <- here(
  "data", "raw",
  "DATA_leaf_d13C_2018_2021_ARBOfun.csv"
)


if (!file.exists(delta_d13C_species_means)) {
  message(
    "The leaf d13C data are currently restricted during peer review. ",
    "If you are a reviewer of this manuscript, please use the data access link ",
    "provided by the editor to download the dataset and place it in ",
    "'data/raw/' before proceeding."
  )
}


```

## 2.3. Stomatal anatomy traits

Kretz, L., Sachsenmaier, L., von Sivers, L., Nabel, N., Wirth, C., & Weigelt, A. (2026). 
Stomatal density and size of 27 angiosperm tree species (ARBOfun research arboretum, Germany) 
(Version 1.0.0) [Dataset]. Zenodo. https://doi.org/10.5281/zenodo.22117766

Sachsenmaier, L., Franz, S., Kahl, A., Kretz, L., & Wirth, C. (2026). 
Stomatal density and size of 12 conifer tree species (1.0.0) 
[Data set]. Zenodo. https://doi.org/10.5281/zenodo.21098915


```{r download stomata anatomy traits}

# Angiosperm stomatal trait data
download.file(
  url = "https://zenodo.org/records/22117766/files/DATA_stomatal_density_size_angiosperm_tree_species_ARBOfun.csv?download=1",
  destfile = here(
    "data", "raw",
    "Kretz_Sachsenmaier_et_al_2026_angiosperm_stomatal_trait_data.csv"
  ),
  mode = "wb"
)


# Gymnosperm stomatal trait data
download.file(
  url = "https://zenodo.org/records/21098915/files/Sachsenmaier_et_al_2026_conifer_stomatal_trait_data.csv?download=1",
  destfile = here(
    "data", "raw",
    "Sachsenmaier_et_al_2026_conifer_stomatal_trait_data.csv"
  ),
  mode = "wb"
)

```

## 2.4. Leaf economic traits

Please note, the data is currently under review at Zenodo, but should be publicly available soon via this DOI:

10.5281/zenodo.21821305 

Please download it and put it into the raw data folder.

```{r downlaod leaf economic traits}

leaf_traits_data_location <- "https://doi.org/10.5281/zenodo.21821305"


```


