---
title: "Stomatal Behaviour 2026 - 01 Data Cleaning Script"
author: "Lena Sachsenmaier"
output:
  html_document:
    df_print: paged
contact: lena.sachsenmaier.research@gmail.com.de
update: "2026-07-30"
updated: "`r format(Sys.time(), '%d %B, %Y')`"
---


Loading packages needed

```{r setup}

library(tidyverse) # basic data manipulation
library(lubridate) # for wrangling with dates and times
library(here) # reproducible file paths

```


### 1. Load the data frame

```{r load data}

data <- read_csv(here("data", "raw", "data_stomatal_conductance.csv"))

```


### 2. Drivers sanity check

based on the equations in the manual certain things must be excluded:
* any value that is lower than VPD = 0.05 kPa
* any value that is higher than RH = 95 %
```{r drivers sanity check}

sanity_check_VPD <- data %>% 
  filter(VPDleaf < 0.05)
# result: all values are fine!

sanity_check_RH <- data %>% 
  filter(rh_s > 95 | rh_r > 95)
# result: all values are fine!


# delete dataframes, not needed anymore
rm(sanity_check_RH, sanity_check_VPD)

```


### 3. Correction of light measurements (via dummy flag)

During the campaign it could have happened, that we measured gsw, but the light sensor
was blocked or shadowed, i.e. the light measurement is not reliable to the corresponding gsw.
In such cases people were told to measure directly afterwards again so that the light sensor is not 
blocked, but gsw is not reliable and flag this value as "dummy_light" in the device's remarks.
For all those values that are flagged with "dummy_light" we need to replace the Qamb value of the measurement
before (same tree, same leaf, same time range) and delete the dummy Qamb values afterwards.


```{r dummy light}

# Dummy light measurements contain a reliable Qamb value measured
# immediately after the original measurement on the same leaf.
# They are used to replace Qamb values when the light sensor
# was temporarily blocked.
data_dummylight <- data %>%
  arrange(date_time) %>%
  # Round all times to nearest 10 minutes
  mutate(nearest_time = round_date(date_time, unit = "10 mins"))

dummy <- data_dummylight %>%
  filter(remarks_device == "dummy_light") %>% 
  mutate(Qamb_dummy = Qamb) %>% 
  # only keep what we need, so that gsw cannot be used of this dataset, its invalid!
  dplyr::select(c(species, tree_id, leaf_nr, nearest_time, Qamb_dummy))

data_with_dummy_qamb <- data_dummylight %>% 
  left_join(dummy, by=c("species", "tree_id", "leaf_nr", "nearest_time"))

# delete dummry data frame, not needed anymore
rm(dummy)

# replace Qamb where dummy exists and mark the source
data_qamb <- data_with_dummy_qamb %>%
  mutate(
    # use Qamb_dummy when it exists, otherwise keep Qamb
    Qamb = coalesce(Qamb_dummy, Qamb), 
    Qamb_source = if_else(!is.na(Qamb_dummy), "dummy", "original")
  ) %>%
  filter(is.na(remarks_device) | remarks_device != "dummy_light") %>%
  dplyr::select(-Qamb_dummy)


```


### 4. Exclusion of immediate double-measurements

Sometimes it might have occurred, that people were surprised by a measurement value 
(unusually high, or unusually low) and directly measured this leaf again, to be sure.
When they inputted the right leaf number again, but did not specify the first value 
as to-be-deleted, the wrong one is still in. But in such cases the value should not 
stay in double, but we only want to keep the second measurement value.
So in short:
If the same leaf was measured twice within a short time window, 
we assume the first measurement was erroneous and only the second measurement is retained.
We define a time threshold of 5 minutes. We removed 10 erroneous measurements with that:

```{r exclude immediate double values}

data_no_repeats <- data_qamb %>%
  arrange(species, tree_id, leaf_nr, date_time) %>%
  group_by(species, tree_id, leaf_nr) %>%
  mutate(
    time_diff = as.numeric(difftime(date_time, lag(date_time), units = "mins")),
    immediate_repeat = !is.na(time_diff) & time_diff <= 5
  ) %>% 
  mutate(
    followed_by_repeat = lead(immediate_repeat, default = FALSE)
  ) %>%
  filter(!followed_by_repeat) %>%
  ungroup()

# We removed 10 measurements

```


### 5. Chamber leakage quality control

All values with very high leakage are not reliable! 
In the needle porometer (LI-600N, LICOR) it's normal to have quite high leakage,
because the needle is going through the chamber and the gaskets cannot always close perfectly.
We set the threshold value to 20 % leakage for conifers (LI-600N, LICOR), 
and to 15 % leakage for broadleaveds (LI-600, LICOR) and every value with higher leakage gets excluded.

```{r leakage, fig.height=2, fig_width=3}

# Inspect leakage distributions separately for broadleaveds and conifers
# to evaluate appropriate quality-control thresholds.
data_no_repeats %>% 
  ggplot() + 
  geom_histogram(aes(leak_pct)) + 
  facet_wrap(~tree_group, scales = "free_y")

# Quantify the consequences of alternative leakage thresholds on data retention
data_no_repeats %>%
  group_by(tree_group) %>%
  summarise(
    pct_over_10 = mean(leak_pct > 10, na.rm = TRUE) * 100,
    pct_over_15 = mean(leak_pct > 15, na.rm = TRUE) * 100,
    pct_over_20 = mean(leak_pct > 20, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )


# check, how many values have high leakage with the given leakage percentages (15 % for angio, 20 % for gymno)
# and how many leaves would remain, when we have to take out the high leakage values
leak_overview <- data_no_repeats %>%
  mutate(leak_flag = case_when(tree_group == "broadleaved" & leak_pct > 15 ~ TRUE,
                               tree_group == "conifer" & leak_pct > 20 ~ TRUE,
                               TRUE ~ FALSE)) %>%
  group_by(species, tree_id, leaf_nr) %>%
  summarise(
    n_total = n(),
    n_high_leak = sum(leak_flag, na.rm = TRUE),
    n_remaining = n_total - n_high_leak,
    below_5_after = n_remaining < 5,
    below_6_after = n_remaining < 6,
    below_7_after = n_remaining < 7,
    .groups = "drop"
  )


#  check how many leaves would remain per species
leak_overview_sp <- leak_overview %>%
  group_by(species) %>%
  summarise(
    n_leaves_total = n(),
    n_leaves_below_5 = sum(below_5_after),
    n_leaves_below_6 = sum(below_6_after),
    n_leaves_below_7 = sum(below_7_after),
    pct_leaves_below_5 = 100 * mean(below_5_after),
    mean_meas_before = mean(n_total),
    mean_meas_after = mean(n_remaining),
    .groups = "drop"
  )

# only keep the leaves that have more minimum 5 measurements per leaf
# (because otherwise it would be difficult to fit a leaf-level curve into these few points)
leaves_to_exclude_leak <- leak_overview %>%
  filter(n_remaining < 5) %>%
  dplyr::select(species, tree_id, leaf_nr)
# result: we need to exclude 7 leaves completely from the data.

# Exclude those 7 leaves:
data_cleaned_leak <- data_no_repeats %>%
  anti_join(leaves_to_exclude_leak,
            by = c("species", "tree_id", "leaf_nr"))


```

### 6. Physiologically unreliable measurements

Any measurement of stomatal conductance (gsw) < 0, can physically not be true and is a measurement error.
However, sometimes it can be just due to the degree of measurement precision, 
that leads to negative gsw, in case gsw is close to zero.

These data can be converted to 0 if it is certain they are occurring due to low conductance 
(i.e., it’s around midday/there’s light yet the PhiPS2 and the Etr measured by the fluorometer are negative). 
Even in those cases where it indeed looks like they are due to low conductance, 
only pass them to 0 if the leakage rate is very low (e.g., <5%).

Otherwise, if there is no definite indication that they are due to a low gs, remove them entirely -- 
the reason is, that you could get a negative read for an actually high gs if the measurement of the flow rates fails. 
For example, a co—occurring negative gsw with positive PhiPS2 and positive Etr are not 
physiologically possible yet do not indicate a very low gsw either – so we decided to remove these data.


```{r negative gsw}

# Negative stomatal conductance values indicate either
# measurement uncertainty around zero or instrument failure.
# Quantify their frequency before applying the filtering rules.
data_cleaned_leak %>%
  summarise(n_negative_gsw = sum(gsw < 0))
# result: 71 observations have gsw < 0

# neg_gsw_sp <- data_cleaned_leak %>% filter(gsw<0) %>% group_by(species, tree_id, leaf_nr) %>% summarize(n=n())

# check, whether are there any species-specific patterns in negative gsw:
data_cleaned_leak %>% 
  filter(gsw<0) %>% 
  count(species)


# filter the unreliable values out:
data_cleaned_physiological <- 
  data_cleaned_leak %>% 
  mutate(
    reliability = case_when(
      tree_group == "broadleaved" & gsw<0 & leak_pct>5 ~ "unreliable",
      tree_group == "conifer" & gsw<0 & leak_pct>10 ~ "unreliable",
      gsw<0 & PhiPS2<0 & ETR<0 & Qamb>50 ~ "zero",
      # in those cases where it indeed looks like they are due to low conductance, we only pass them to 0 if the leakage rate is very low (<5%).
      tree_group == "broadleaved" & gsw<0 & leak_pct<5 ~ "zero",
      tree_group == "conifer" & gsw<0 & leak_pct<10 ~ "zero",
      gsw>= 0 ~ "reliable",
      TRUE ~ NA_character_
    )) %>% 
      # remove the ones that are not reliable
  filter(reliability!="unreliable") %>% 
      # convert the ones to zero that might be real measurements, but are just negative due to the device inaccuracy
  mutate(
      gsw = case_when(reliability=="zero" ~ 0,
                      reliability=="reliable" ~ gsw,
                      TRUE ~ gsw)
      )

# check what changed, with a summary table
data_cleaned_physiological %>% 
  group_by(tree_group, reliability) %>% 
  summarize(
    `Count` = n(),
    .groups = "drop"
  )

```



### 7. Exclusion of leaves with less than 5 data points

```{r quick test how many measurement we have per tree}

leaves_with_too_little_data <- 
  data_cleaned_physiological %>% 
  group_by(tree_id, date, leaf_nr) %>% 
  summarize(datapoints_per_leaf = n(), .groups="drop") %>% 
  filter(datapoints_per_leaf < 5) %>%
  dplyr::select(tree_id, date, leaf_nr)

# exclude the leaves that have now too little data points (below 5)
data_final <- 
  data_cleaned_physiological %>%
  anti_join(leaves_with_too_little_data,
            by = c("tree_id", "date", "leaf_nr"))
# 16 measurements are excluded with this step

```


### 8. Summary of the final data
```{r check of the final dataframe}

# Measurements per tree
tree_summary <- data_final %>%
  group_by(species, tree_id) %>%
  summarise(
    measurements = n(),
    leaves = n_distinct(leaf_nr),
    .groups = "drop"
  )

# Species-level overview
final_dataset_summary <- 
  tree_summary %>%
  group_by(species) %>%
  summarise(
    trees = n(),
    measurements_per_tree_min = min(measurements),
    measurements_per_tree_max = max(measurements),
    measurements = sum(measurements),
    leaves_per_tree_min = min(leaves),
    leaves_per_tree_max = max(leaves),
    leaves = sum(leaves),
    .groups = "drop"
  )

final_dataset_summary

```

### 9. Export of the final data

```{r write the dataframe}

write_csv(data_final, here("data", "processed", "data_cleaned.csv"))

```