---
title: "Stomatal Behaviour 2026 - Main model and metrics"
author: "Lena Sachsenmaier"
output:
  html_document:
    df_print: paged
contact: lena.sachsenmaier.research@gmail.de
created: "2025-07-30"
updated: "`r format(Sys.time(), '%d %B, %Y')`"
---



```{r setup}
library(tidyverse) # core data manipulation
library(here) # reproducible file paths

# Bayesian modelling
library(brms)
library(loo)
library(bayesplot) # for diagnostics and figures of brms() models
library(tidybayes) # for easy viusalization of bayesian models
library(posterior)

# correlation analysis
library(Hmisc) # for correlation matrix
library(scales) # for rescaling variables for PCA

# moving window calculations
library(zoo)

# figure layout
library(patchwork)
library(ggrepel)
library(ggh4x) # for plotting figures (facet with different axis labels)
library(ggnewscale) # for adding a new colour scale

# export supplementary tables
library(writexl) # for writing excel tables

```

# 1.Preparation of data frame

## 1.1 Load data frame

(This data frame is already cleaned by the previous R script 01_cleaning.rmd.)

```{r load data}

data <- read_csv(here("data", "processed", "data_cleaned.csv"))

```


## 1.2. Preparation of variables for modeling

```{r variable preparation}

data <- data %>% 
  mutate(Qamb_sc = as.numeric(scale(Qamb)),
         species=as.factor(species),
         tree_id=as.factor(tree_id)) %>% 
  # create a unique leaf_id variable
  unite("leaf_id", tree_id, leaf_nr, sep="_", remove=F) %>% 
  mutate(leaf_id = as.factor(leaf_id))

```

## 1.3. Helper vectors and functions

```{r helper functions}

# create grids for later predictions
# for figures it's enough if VPDleaf reaches until 6 kPa
# for calculation of metrics we use a larger range

VPD_grid_plot <- seq(0,6, length.out = 1000)

VPD_grid_metrics <- seq(0,15, length.out = 1000)


create_prediction_grid <- function(sp, VPD_grid) {
  tibble(
    VPDleaf = VPD_grid,
    species = sp,
    tree_id = NA,
    leaf_id = NA,
    # Ambient light is held at its mean observed level.
   # Qamb_sc = 0 corresponds to mean Qamb because Qamb was standardized.
    Qamb_sc = 0
  )
}

# Create prediction data for every measured leaf of one species.
# Leaf and tree identifiers are retained so random effects are included.
create_prediction_grid_leaf <- function(sp, VPD_grid, data) {
  data_species <- data %>% filter(species == sp)
  tibble(
    VPDleaf = rep(VPD_grid, times = nrow(data_species)),
    species = sp,
    tree_id = rep(data_species$tree_id, each = length(VPD_grid)),
    leaf_id = rep(data_species$leaf_id, each = length(VPD_grid)),
    Qamb_sc = 0
  )
}




# Extract the posterior predictions
# Each posterior draw represents one plausible stomatal response curve.
predict_posterior_curves <- function(sp, nd, model, ndraws = 1000) {
  ep <- posterior_epred(
    model,
    newdata = nd,
    # ignore random effects
    re_formula = NA,   # ignore random effects
    ndraws = ndraws
  )
  
  tibble(
    VPD  = rep(nd$VPDleaf, times = ndraws),
    draw = rep(seq_len(ndraws), each = length(nd$VPDleaf)),
    gsw  = as.vector(t(ep))
  )
}


# Calculate the posterior mean prediction per species.
# This represents the average population-level response curve.
predict_mean_curve <- function(sp, nd, model) {
  ep_mean <- posterior_epred(
    model,
    newdata = nd,
    # ignore random effects
    re_formula = NA
  ) %>% colMeans()
  
  tibble(
    VPD = nd$VPDleaf,
    gsw_mean = ep_mean
  )
}



# Predict leaf-specific response curves including random effects.
# Posterior predictions are averaged across sampled draws.
predict_leaf_curves <- function(nd, model, ndraws = 50) {
  ep <- posterior_epred(
    model,
    newdata = nd,
    re_formula = NULL,  # include random effects
    ndraws = ndraws
  )
  # posterior_epred returns n_draws × nrow(nd), need to transpose
  ep_mean <- colMeans(ep)  # mean across draws for each row
  
  tibble(
    VPD = nd$VPDleaf,
    gsw = ep_mean,
    leaf_id = nd$leaf_id,
    tree_id = nd$tree_id
  )
}

# specify colour selection for later figures
pastel <- c("#4EA8DE", "#A29BFE", "#5B7065")

```





# 2. Modelling gsw - Bayesian Framework

## 2.1. Final model

```{r final model}

# the model formula
model_formula <- bf(
  gsw ~ (a * pow(VPDleaf, k) * exp(-b * VPDleaf)),
  a ~ 0 + species + species:Qamb_sc + (1|tree_id/leaf_id),
  b ~ 0 + species + (1|tree_id/leaf_id),
  k ~ 0 + species + (1|tree_id/leaf_id),
  nl = TRUE
)

# priors for the model
priors_model <- c(
  # restricted to be positive
  prior(lognormal(log(0.7), 0.7), nlpar = "a", lb=0),
  prior(lognormal(log(1), 0.7), nlpar = "b", lb=0),
  # restricted to be positive and restricted to not explode, because it's the most sensible parameter
  prior(lognormal(0, 1), nlpar = "k", lb=0, ub=4), 
  prior(student_t(3, 0, 1), class = "sigma"),
  prior(gamma(2, 0.1), class = "nu"))


# fit model (caution: takes several hours)
model <- brm(
  formula = model_formula,
  data = data,
  family = student(),
  prior = priors_model,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  control = list(adapt_delta = 0.95),
  save_pars = save_pars(all = TRUE),
  file = here::here("output", "models", "final_model"),
  file_refit = "on_change"
)

# # Optional:
# # # Reload previously fitted model, in case you did not have time to run it
# model <- readRDS(here::here("output", "models", "final_model.rds"))


```

## 2.2. Posterior predictions
```{r posterior predictions of the model}


species_list <- unique(data$species)
species_list

# predict 100 single posterior draws per species
posterior_draws <- map_df(species_list, function(sp) {
  nd <- create_prediction_grid(sp, VPD_grid_plot)
  predict_posterior_curves(sp, nd, model, ndraws = 100) %>% 
    mutate(species = sp)
})


# predict species mean curves across all draws:
posterior_means <- map_df(species_list, function(sp) {
  nd <- create_prediction_grid(sp, VPD_grid_plot)
  predict_mean_curve(sp, nd, model) %>% 
    mutate(species = sp)
})


```


## 2.3. Model diagnostics

```{r diagnostics final model}

# checking the model summary
summary_model <- as.data.frame(posterior_summary(model))

summary(model)

pp_check(model)

bayes_R2(model)

# checking the residuals

res <- residuals(model)
fit <- fitted(model)

res_vals <- res[, "Estimate"]
fit_vals <- fit[, "Estimate"]


plot(fit_vals, res_vals,
     xlab = "Fitted values",
     ylab = "Residuals")
abline(h = 0, col = "red")

# residuals vs. VPDleaf (my key predictor)
plot(data$VPDleaf, res_vals,
     xlab = "VPDleaf",
     ylab = "Residuals")
abline(h = 0, col = "red")

# distribution of residuals
hist(res_vals, breaks = 30,
     main = "Residual distribution",
     xlab = "Residuals")

## Priors check

model_check_priors <- brm(
  formula = model_formula,
  data = data,
  family = student(),
  prior = priors_model,
  chains = 4,
  cores = 4,
  iter = 4000,
  warmup = 2000,
  sample_prior = "only",
  control = list(adapt_delta = 0.95),
  save_pars = save_pars(all = TRUE),
  file = here::here("output", "model_prior_check"),
  file_refit = "on_change"
)

# # Optional:
# # Reload previously fitted model, in case you did not have time to run it
# model_check_priors <- readRDS(here::here("output", "models", "model_prior_check.rds"))

VPD_grid <- seq(0, 6, length.out = 100)

baseline <- data.frame(
  VPDleaf = VPD_grid,
  species = levels(data$species)[1],
  Qamb_sc = 0
)

# Extract predictor draws
prior_samples_grid <- posterior_epred(
  model_check_priors,
  newdata = baseline,
  transform = TRUE,
  re_formula = NA
)

# Sample 100 draws and format for ggplot
set.seed(123)

draw_indices <- sample(1:nrow(prior_samples_grid), 100)
pp_curves_sampled <- prior_samples_grid[draw_indices, ]

# Convert the matrix to a long-format data frame for ggplot
plot_data <- as.data.frame(t(pp_curves_sampled)) %>%
  mutate(VPDleaf = VPD_grid) %>%
  pivot_longer(
    cols = -VPDleaf,
    names_to = "draw",
    values_to = "gsw"
  )

fig_prior_predictive <- ggplot(plot_data, aes(x = VPDleaf, y = gsw, group = draw)) +
  geom_line(alpha = 0.3, color = "royalblue", linewidth = 0.5) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(
    x = expression(Vapor~Pressure~Deficit~(VPD[leaf]*", kPa")),
    y = expression(Stomatal~Conductance~(italic(g)[sw]*", mol "~m^{-2}~s^{-1}))
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    axis.title = element_text(face = "plain")
  )

## Model convergence - visualization
## Creates the default brms diagnostic plots, including
## trace plots and posterior density plots for all parameters.

diagn_plot_model_final <- plot(model, ask = FALSE)

for (i in seq_along(diagn_plot_model_final)) {
  ggsave(
    filename = here("output", paste0("trace_plot_", i, ".png")),
    plot = diagn_plot_model_final[[i]],
    width = 8,
    height = 6,
    dpi = 300
  )
}

## Check the parameter correlations

post <- as_draws_df(model)

species_list <- unique(data$species)

for (sp in species_list) {

  vars_sp <- c(
    paste0("b_a_species", sp),
    paste0("b_b_species", sp),
    paste0("b_k_species", sp)
  )

  p <- mcmc_pairs(
    post,
    pars = vars_sp
  )

  ggsave(
    filename = here("output", "pairs_species", paste0("pairs_", sp, ".png")),
    plot = p,
    width = 6,
    height = 6
  )
}


```


## 2.3. Model Output - Tab. S4-S5

```{r tables supplement}

# summary of fixed effects
fixed_effects_long <- as.data.frame(posterior_summary(model)) %>%  
  rownames_to_column(var = "Parameter_Raw") %>% 
  filter(str_detect(Parameter_Raw, "^b_(a|b|k)_species")) %>% 
  mutate(
    Effect_Type = "Fixed",
    # Clean parameter names (e.g., "a", "a:Qamb_sc")
    Parameter = str_replace(Parameter_Raw, "^b_", "") %>% 
                str_replace("_species[A-Za-z_]+", ""),
    # Extract species name
    Species = str_extract(Parameter_Raw, "(?<=_species)[A-Za-z_]+")
  )

# summary of random effects:
random_effects_long <- as.data.frame(posterior_summary(model)) %>%  
  rownames_to_column(var = "Parameter_Raw") %>% 
  filter(str_detect(Parameter_Raw, "^sd_|^sigma|^nu")) %>%
  mutate(
    Species = NA_character_, 
    Effect_Type = case_when(
      str_detect(Parameter_Raw, "tree_id:leaf_id") ~ "Random: tree_id:leaf_id",
      str_detect(Parameter_Raw, "tree_id") ~ "Random: tree_id",
      str_detect(Parameter_Raw, "sigma")   ~ "Residual (sigma)",
      str_detect(Parameter_Raw, "nu")      ~ "Distribution Parameter (nu)",
      TRUE                                 ~ "Random: Group"
    ),
    Parameter = case_when(
      str_detect(Parameter_Raw, "sigma") ~ "Sigma",
      str_detect(Parameter_Raw, "nu")    ~ "Nu",
      str_detect(Parameter_Raw, "__")    ~ {
        raw_param <- str_extract(Parameter_Raw, "(?<=__)[a-zA-Z_]+")
        # Remove "Intercept" if it's attached to a specific parameter like "a_Intercept"
        clean_param <- str_replace(raw_param, "_Intercept|^Intercept", "")
        # If it was just the bare intercept, name it "Intercept", otherwise name it "SD_a", "SD_b", etc.
        if_else(clean_param == "" | is.na(clean_param), "SD_Intercept", paste0("SD_", clean_param))
      },
      TRUE                                ~ "SD_Global"
    )
  ) %>% 
  mutate(Parameter = case_when(Parameter=="SD_a" ~ "a",
                               Parameter=="SD_b" ~ "b",
                               Parameter=="SD_k" ~ "k",
                               TRUE ~ Parameter))


# join the fixed and random effects summary
summary_table <- bind_rows(fixed_effects_long, random_effects_long) %>% 
  mutate(
    Cell_Value = paste0(round(Estimate, 2), " [", round(`Q2.5`, 2), ", ", round(`Q97.5`, 2), "]")
  ) %>% 
  dplyr::select(Effect_Type, Species, Parameter, Cell_Value) %>% 
  pivot_wider(
    names_from = Parameter,
    values_from = Cell_Value
  )

summary_table





# Find the highest Rhat in the entire model
max_rhat <- max(brms::rhat(model), na.rm = TRUE)

# Find the lowest Effective Sample Sizes in the entire model
min_bulk <- min(brms::neff_ratio(model) * brms::nsamples(model), na.rm = TRUE)

cat("Worst Rhat in the entire model:", round(max_rhat, 3), "\n")
cat("Lowest Bulk-ESS in the entire model:", round(min_bulk, 0), "\n")





diag_df <- brms::as_draws_array(model) %>%
  summarise_draws("rhat", "ess_bulk", "ess_tail") %>%
  rename(
    Parameter_Raw = variable,
    Rhat = rhat,
    Bulk_ESS = ess_bulk,
    Tail_ESS = ess_tail
  )

all_diagnostics <- diag_df %>%
  dplyr::select(Parameter_Raw, Rhat, Bulk_ESS, Tail_ESS)

# all_diagnostics <- as.data.frame(posterior_summary(model)) %>% 
#   rownames_to_column(var = "Parameter_Raw") %>%
#   dplyr::select(Parameter_Raw, Rhat, Bulk_ESS, Tail_ESS)

# Keep fixed effects AND the global random / residual / nu structures
diagnostics_long <- all_diagnostics %>% 
  filter(
    str_detect(Parameter_Raw, "^b_(a|b|k)_species") | # Fixed species parameters
    str_detect(Parameter_Raw, "^sd_|^sigma|^nu")      # Global random, residual, and nu
  ) %>% 
  mutate(
    # Determine effect type
    Effect_Type = case_when(
      str_detect(Parameter_Raw, "^b_")             ~ "Fixed",
      str_detect(Parameter_Raw, "tree_id:leaf_id") ~ "Random: tree_id:leaf_id",
      str_detect(Parameter_Raw, "tree_id")         ~ "Random: tree_id",
      str_detect(Parameter_Raw, "sigma")           ~ "Residual (sigma)",
      str_detect(Parameter_Raw, "nu")              ~ "Distribution Parameter (nu)",
      TRUE                                         ~ "Random: Group"
    ),
    
    # Extract species name only for fixed rows
    Species = if_else(Effect_Type == "Fixed", 
                      str_extract(Parameter_Raw, "(?<=_species)[A-Za-z_]+"), 
                      NA_character_),
    
    # Clean the Parameter column so everything aligns horizontally
    Parameter = case_when(
      Effect_Type != "Fixed" & str_detect(Parameter_Raw, "sigma") ~ "Sigma",
      Effect_Type != "Fixed" & str_detect(Parameter_Raw, "nu")    ~ "Nu",
      str_detect(Parameter_Raw, "__") ~ {
        raw_param <- str_extract(Parameter_Raw, "(?<=__)[a-zA-Z_]+")
        clean_param <- str_replace(raw_param, "_Intercept|^Intercept", "")
        if_else(clean_param == "" | is.na(clean_param), "Intercept", clean_param)
      },
      TRUE ~ str_replace(Parameter_Raw, "^b_", "") %>% str_replace("_species[A-Za-z_]+", "")
    )
  ) %>% 
  dplyr::select(Parameter, Species, Effect_Type, Rhat, Bulk_ESS, Tail_ESS)


wide_diagnostics_columns <- diagnostics_long %>%
  pivot_wider(
    names_from = Parameter,
    values_from = c(Rhat, Bulk_ESS, Tail_ESS),
    names_glue = "{Parameter}_{.value}"
  ) %>% 
  # Round numbers nicely
  mutate(across(contains("Rhat"), ~ sprintf("%.2f", .x))) %>% 
  mutate(across(contains("ESS"), ~ sprintf("%.0f", .x)))

wide_diagnostics_columns

# wide_diagnostics_columns <- wide_diagnostics_columns %>%  mutate(Species= case_when(
#                                   Species== "Pla_ace" ~ "Pla_his",
#                                   Species=="Sor_dom" ~ "Cor_dom",
#                                   Species== "Sor_tor" ~ "Tor_gla",
#                                   TRUE ~ Species)) %>% 
#   arrange(Species)


# export the supplementary tables S4 and S5

write_xlsx(summary_table, here("output", "Table_S5_model_summary_table.xlsx"))

write_xlsx(wide_diagnostics_columns, here("output", "Table_S4_model_diagnostics_table.xlsx"))

```



# 3. Stomatal Behaviour Metrics

## 3.1. Extraction of metrics

```{r extract metrics from the posterior curves}


# Create prediction data for all species using the metrics VPD range
newdata <- map_df(
  unique(data$species),
  ~ create_prediction_grid(.x, VPD_grid_metrics)
)

# Population-level posterior predictions
ep <- posterior_epred(
  model,
  newdata = newdata,
  re_formula = NA
)

# here we use all the available draws (n=4000)
ndraws <- nrow(ep)


# Convert to tidy format

posterior_curves <- tibble(
  species = rep(newdata$species, times = ndraws),
  VPDleaf = rep(newdata$VPDleaf, times = ndraws),
  draw = rep(seq_len(ndraws), each = nrow(newdata)),
  gsw = as.vector(t(ep))
) %>% 
  arrange(species, draw, VPDleaf) %>%
  group_by(species, draw)


# extraction of metrics
stomatal_behaviour_metrics <- posterior_curves %>%
  group_modify(~{

    vpd   <- .x$VPDleaf
    epred <- .x$gsw

    ord <- order(vpd)

    vpd   <- vpd[ord]
    epred <- epred[ord]

    n <- length(vpd)

    # safety
    if (n < 6) {
      return(tibble(
        VPD_opt      = NA_real_,
        gsw_max      = NA_real_,
        VPD50         = NA_real_,
        Sclosure = NA_real_
      ))
    }

    # find the peak of the curve
    # and extract the peak value (gs_max) and the VPDopt

    peak_idx <- which.max(epred)

    VPD_opt <- vpd[peak_idx]
    gs_max <- epred[peak_idx]

    half_val <- 0.5 * gs_max

    # extract VPD50

    VPD50 <- NA_real_

    if (peak_idx < n) {

      idx_after_peak <- seq(peak_idx + 1, n)

      idx_half <- idx_after_peak[
        which.min(abs(epred[idx_after_peak] - half_val))
      ]

      VPD50 <- vpd[idx_half]
    }

    # smoothing for the relative curve
    # so that the maximum slope is not confounded
    
    epred_smooth <- zoo::rollmean(
      epred,
      k = 5,
      fill = "extend"
    )

    # curve standardized by its gs max
    # so that later the slopes are comparable
    epred_rel <- epred / gs_max

    epred_rel_smooth <- zoo::rollmean(
      epred_rel,
      k = 5,
      fill = "extend"
    )

    # first derivates

    slope <- diff(epred_smooth) / diff(vpd)

    slope_rel <- diff(epred_rel_smooth) / diff(vpd)

    # midpoint slope
    vpd_slope <- (vpd[-1] + vpd[-n]) / 2

    # declining limb only
    keep_decline <- vpd_slope >= VPD_opt

    slope_after_peak     <- slope[keep_decline]
    slope_rel_after_peak <- slope_rel[keep_decline]

    vpd_after_peak <- vpd_slope[keep_decline]

    # strongest decline
    min_slope <- min(
      slope_after_peak,
      na.rm = TRUE
    )
    # make the value absolute to be able to compare it as "closure rate"
    Sclosure <- abs(min(
      slope_rel_after_peak,
      na.rm = TRUE
    ))

    # output table

    tibble(
      VPD_opt      = VPD_opt,
      gs_max      = gs_max,
      VPD50         = VPD50,
      Sclosure = Sclosure
    )

  })



# join the old speices names

stomatal_behaviour_metrics <- 
  stomatal_behaviour_metrics %>%
  mutate(species = case_when(species=="Pla_ace" ~ "Pla_his", 
                            species=="Sor_dom" ~ "Cor_dom",
                            species=="Sor_tor" ~ "Tor_gla",
                            TRUE ~ species))



# the mean of all draws per species

stomatal_behaviour_species_metrics <- 
  stomatal_behaviour_metrics %>%
  group_by(species) %>%
  summarise(

    gs_max_l95  = quantile(gs_max, 0.025, na.rm = TRUE),
    gs_max_u95  = quantile(gs_max, 0.975, na.rm = TRUE),
    gs_max = mean(gs_max, na.rm = TRUE),
    

    VPD_opt_l95  = quantile(VPD_opt, 0.025, na.rm = TRUE),
    VPD_opt_u95  = quantile(VPD_opt, 0.975, na.rm = TRUE),
    VPD_opt = mean(VPD_opt, na.rm = TRUE),


    VPD50_l95  = quantile(VPD50, 0.025, na.rm = TRUE),
    VPD50_u95  = quantile(VPD50, 0.975, na.rm = TRUE),
    VPD50 = mean(VPD50, na.rm = TRUE),


    Sclosure_l95  = quantile(Sclosure, 0.025, na.rm = TRUE),
    Sclosure_u95  = quantile(Sclosure, 0.975, na.rm = TRUE),
    Sclosure = mean(Sclosure, na.rm = TRUE),
  )

# join the status conifer or broadleaved to the data

tree_group_column <- data %>% dplyr::select(species, tree_group) %>% distinct()

stomatal_behaviour_species_metrics <- 
  stomatal_behaviour_species_metrics %>%
  left_join(tree_group_column)




write_csv(stomatal_behaviour_metrics, here::here("data", "processed", "stomatal_behaviour_metrics.csv"))
write_csv(stomatal_behaviour_species_metrics, here::here("data", "processed", "stomatal_behaviour_species_metrics.csv"))


```




## 3.3. Metrics overview - Table S6
```{r supplement table S6}

# 1. Calculate the raw statistics and format into single cells
metrics_formatted_table <- stomatal_behaviour_species_metrics %>%

  mutate(
    gs_max_formatted = sprintf("%.3f [%.3f, %.3f]", gs_max, gs_max_l95, gs_max_u95),
    VPD_opt_formatted = sprintf("%.2f [%.2f, %.2f]", VPD_opt, VPD_opt_l95, VPD_opt_u95),
    VPD50_formatted   = sprintf("%.2f [%.2f, %.2f]", VPD50, VPD50_l95, VPD50_u95),
    Sclosure_formatted = sprintf("%.2f [%.2f, %.2f]", Sclosure, Sclosure_l95, Sclosure_u95)
  ) %>%
  
  dplyr::select(
    Species = species,
    `gs_max (mol m-2 s-1)` = gs_max_formatted,
    `VPD_opt (kPa)`        = VPD_opt_formatted,
    `VPD50 (kPa)`          = VPD50_formatted,
    `Sclosure`             = Sclosure_formatted
  )

metrics_formatted_table

write_xlsx(metrics_formatted_table, here::here("output", "Tab_S6_stomatal_behaviour_species_metrics.xlsx"))

```

##3.4. Differences gymno vs. angio - Fig. S8

```{r gymno vs. angio}

# Join  the tree group onto the draw-level data
stomatal_behaviour_metrics_with_clade <- 
  stomatal_behaviour_metrics %>% 
  left_join(tree_group_column)

# For each draw, compute group means, then the difference
draw_diffs <- stomatal_behaviour_metrics_with_clade %>%
  pivot_longer(cols = c(VPD_opt, gs_max, VPD50, Sclosure),
               names_to = "metric", values_to = "value") %>%
  group_by(draw, metric, tree_group) %>%
  summarise(group_mean = mean(value), .groups = "drop") %>%
  pivot_wider(names_from = tree_group, values_from = group_mean) %>%
  mutate(diff = gymnosperm - angiosperm)

# Summarise the posterior of the difference, per metric
diff_summary_table <- draw_diffs %>%
  group_by(metric) %>%
  summarise(
    mean_diff   = mean(diff),
    lower95     = quantile(diff, 0.025),
    upper95     = quantile(diff, 0.975),
    prob_positive = mean(diff > 0),          # P(angiosperm > gymnosperm)
    prob_direction = pmax(prob_positive, 1 - prob_positive)
  )



# Figure
# Get max absolute difference per metric, so we can build symmetric limits
lims <- draw_diffs %>%
  group_by(metric) %>%
  summarise(max_abs = max(abs(diff)) * 1.05)  # 5% padding


# Create invisible "anchor" points at +/- max_abs per metric
#    geom_blank will use these to set symmetric x-limits per facet
anchor_df <- lims %>%
  tidyr::pivot_longer(max_abs, values_to = "x") %>%
  mutate(x = list(c(-1, 1))) %>%
  tidyr::unnest(x) %>%
  mutate(x = x * (lims$max_abs[match(metric, lims$metric)]))


metric_order <- c("gs_max", "VPD_opt", "VPD50", "Sclosure")

metric_labels <- c(
  gs_max   = "atop(g[max], (mol~m^{-2}~s^{-1}))",
  VPD_opt  = "atop(VPD[opt], (kPa))",
  VPD50    = "atop(VPD[50], (kPa))",
  Sclosure = "atop(S[closure], (mol~m^{-2}~s^{-1}~kPa^{-1}))"
)

group_post <- draw_diffs %>%
  dplyr::select(draw, metric, angiosperm, gymnosperm) %>%
  pivot_longer(cols = c(angiosperm, gymnosperm),
               names_to = "group", values_to = "value") %>%
  mutate(
    group = recode(group, angiosperm = "broadleaved", gymnosperm = "conifer"),
    metric = factor(metric, levels = metric_order)
  )

# Build significance summary per metric (from draw_diffs, computed earlier) --
sig_table <- draw_diffs %>%
  group_by(metric) %>%
  summarise(
    mean_diff = mean(diff),
    lower95   = quantile(diff, 0.025),
    upper95   = quantile(diff, 0.975),
    .groups = "drop"
  ) %>%
  mutate(
    sig_label = if_else(lower95 > 0 | upper95 < 0, "*", ""),
    label_text = sprintf("Δ = %.2f [%.2f, %.2f] %s",
                          mean_diff, lower95, upper95, sig_label))
    
label_pos <- group_post %>%
  group_by(metric) %>%
  summarise(x_pos = mean(range(value)), .groups = "drop") %>%
  left_join(sig_table, by = "metric") %>%
  mutate(metric = factor(metric, levels = metric_order))


fig_metric_diff_tree_group <- ggplot(group_post, aes(x = value, fill = group)) +
  geom_histogram(position = "identity", bins = 60, alpha = 0.5, color = NA) +
  geom_text(data = label_pos,
            aes(x = x_pos, y = Inf, label = label_text),
            inherit.aes = FALSE,
            vjust = 1.5, size = 3.2, fontface = "bold") +
  facet_wrap(~ metric, scales = "free", nrow = 1,
             labeller = as_labeller(metric_labels, label_parsed)) +
  scale_fill_manual(values = c(broadleaved = "#41624D", conifer = "#7798B9")) +
  labs(x = "Posterior group mean", y = "Count", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", 
        strip.text = element_text(face = "bold"))

ggsave(fig_metric_diff_tree_group, here("output", "Fig_S8_clade_differences.png", height=6, width=10))

```


# 4. Figures (Main)

## 4.1. Fig. 1ab - Modeling example

```{r figures example Sor_auc}

# Just choose one example species, here Sorbus aucuparia
example_species <- "Sor_auc"
data_target <- data %>% filter(species == example_species)

set.seed(321) # Ensure consistency with your previous draws


nd_full <- create_prediction_grid(example_species, VPD_grid_plot)
draws_df <- predict_posterior_curves(example_species, nd_full, model, ndraws = 300)
mean_df  <- predict_mean_curve(example_species, nd_full, model)

p_full <- ggplot() +
  geom_point(
    data = data_target,
    aes(VPDleaf, gsw),
    alpha = 0.2,
    size = 1.6
  ) +
geom_line(
    data = draws_df, 
    aes(VPD, gsw, group = draw), 
    alpha = 0.2,
    linewidth = 0.25,
    color = "grey"
  ) +
  geom_line(
    data = mean_df,
    aes(VPD, gsw_mean),
    linewidth = 1.2
  ) +
  labs(
    x = expression("Leaf-to-air Vapor Pressure Deficit, VPD"["L"] * " (kPa)"),
    y = expression("Stomatal Conductance, g"["s"] * " (mol m"^-2 * " s"^-1 * ")")
  ) +
  ylim(0,0.61) +
  theme_classic(base_size=18)


# Figure on the leaf_id level

nd_leaf <- create_prediction_grid_leaf(example_species, VPD_grid_plot, data_target)
mean_df_leaf <- predict_leaf_curves(nd_leaf, model, ndraws = 50)

data_target <- data_target %>%
  group_by(tree_id) %>%
  mutate(leaf_index = as.factor(match(leaf_id, unique(leaf_id)))) %>%
  ungroup()

mean_df_leaf <- mean_df_leaf %>%
  group_by(tree_id) %>%
  mutate(leaf_index = as.factor(match(leaf_id, unique(leaf_id)))) %>%
  ungroup()


p_leaf <- ggplot() +
  geom_point(
    data = data_target,
    aes(VPDleaf, gsw, colour = tree_id, alpha = leaf_index),
    size = 1.6
  ) +
  geom_line(
    data = mean_df_leaf,
    aes(VPD, gsw, group = leaf_id, color = tree_id, alpha=leaf_index),
    linewidth = 0.4
  ) +
  scale_color_manual(values = pastel,
                     name = "Tree Individual",
                     labels = c("1", "2", "3")) +
  scale_alpha_manual(
    values = c(
  "1" = 0.25, 
  "2" = 0.35, 
  "3" = 0.48, 
  "4" = 0.60, 
  "5" = 0.73, 
  "6" = 0.85, 
  "7" = 1.00
),
    name = "Leaf Individual",
    labels = paste(1:7)
  ) +
  labs(
    x = expression("Leaf-to-air Vapor Pressure Deficit, VPD"["L"] * " (kPa)"),
    y = expression("Stomatal Conductance, g"["s"] * " (mol m"^-2 * " s"^-1 * ")")
  ) +
  xlim(0, 6) +
    ylim(0,0.61) +
  theme_classic(base_size=18) +
theme(
    # Move the legend inside the plot area 
    legend.position = c(0.98, 0.82),
    # Anchor the legend box by its top-right corner
    legend.justification = c(1, 1),  

    legend.background = element_rect(fill = "transparent", color = NA),
    legend.box.background = element_rect(fill = "transparent", color = NA),

    legend.title = element_text(face = "bold", size = 9),
    legend.text = element_text(size = 8),
    legend.spacing.y = unit(0.1, "cm"),
    
    legend.key.size = unit(0.4, "cm") 
  )


combined_plot <- (p_full + p_leaf) + 
  plot_annotation(
    tag_levels = list(c("(a)", "(b)"))
  ) & 
  theme(
    plot.tag = element_text(face = "bold", size = 12)
  )


```

## 4.2. Fig. 1c - Extraction of metrics

```{r figure metric extraction}

# Metrics are extracted here from the posterior mean curve only
# to illustrate the calculation procedure.
# Quantitative analyses use posterior distributions from stomatal_behaviour_metrics.

species_df <- posterior_means %>% filter(species == example_species)

# Find the peak directly on the black line plotted
gs_max_val <- max(species_df$gsw_mean, na.rm = TRUE)
vpd_opt    <- species_df$VPD[which.max(species_df$gsw_mean)]

# Find the 50% drop directly on the black line plotted
gs_50_val    <- gs_max_val * 0.50
post_peak_df <- species_df %>% filter(VPD >= vpd_opt)
vpd_50       <- post_peak_df$VPD[which.min(abs(post_peak_df$gsw_mean - gs_50_val))]

# Calculate the absolute slope geometry for the tangent line
abs_slopes    <- diff(post_peak_df$gsw_mean) / diff(post_peak_df$VPD)
min_slope_idx <- which.min(abs_slopes)

m_slope_abs   <- abs_slopes[min_slope_idx]
vpd_Sclosure  <- post_peak_df$VPD[min_slope_idx]
gs_Sclosure   <- post_peak_df$gsw_mean[min_slope_idx]
intercept_val <- gs_Sclosure - (m_slope_abs * vpd_Sclosure)



fig_extraction_method <- 
  ggplot(species_df, aes(VPD, gsw_mean)) +
  geom_line(linewidth = 1.1, color = "black") +
  geom_abline(intercept = intercept_val, slope = m_slope_abs, color = "#059669", 
              linetype = "solid", linewidth = 0.8) +
  geom_segment(aes(x = vpd_opt, y = 0, xend = vpd_opt, yend = gs_max_val), 
               linetype = "dashed", color = "#9333EA") +
  geom_segment(aes(x = 0, y = gs_max_val, xend = vpd_opt, yend = gs_max_val), 
               linetype = "dashed", color = "#3B5277") +
  geom_segment(aes(x = 0, y = gs_50_val, xend = vpd_50, yend = gs_50_val), 
               linetype = "twodash", color = "#C2410C") +
  geom_segment(aes(x = vpd_50, y = 0, xend = vpd_50, yend = gs_50_val), 
               linetype = "twodash", color = "#C2410C") +
  annotate("text", x = 0.1, y = gs_max_val * 1.03, 
           label = "bold(g)[bold(max)]", parse = TRUE, hjust = 0, colour = "#3B5277") +
  annotate("text", x = vpd_opt, y = gs_max_val * 0.10, 
           label = "bold(VPD)[bold(opt)]", parse = TRUE, vjust = -0.5, angle = 90, colour = "#9333EA") +
  annotate("text", x = vpd_50, y = gs_max_val * 0.10, 
           label = "bold(VPD)[bold(50)]", parse = TRUE, vjust = -0.5, angle = 90, color = "#C2410C") +
  annotate("text", x = vpd_Sclosure + 0.2, y = gs_Sclosure + 0.02, 
           label = "bold(S)[bold(closure)]", parse = TRUE, color = "#059669") +

  labs(
    x = expression("Leaf-to-air Vapor Pressure Deficit, VPD"["L"] * " (kPa)"),
    y = expression("Stomatal Conductance, g"["s"] * " (mol m"^-2 * " s"^-1 * ")")
  ) +
  
  scale_x_continuous(expand = c(0, 0), limits = c(0, max(species_df$VPD) * 1.05)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, gs_max_val * 1.1)) +
  theme_classic(base_size=18)

fig_extraction_method


combined_fig_1abc <- (p_full + p_leaf) / fig_extraction_method +
  plot_layout(heights = c(1, 1.3)) +  # give more vertical room to panel c if needed
  plot_layout(widths =  c(1.3, 1)) +
  plot_annotation(
    tag_levels = list(c("(a)", "(b)", "(c)"))
  ) &
  theme(
    plot.tag = element_text(face = "bold", size = 18)
  )

ggsave(
  filename = here::here("output", "Figure_1abc.png"),
  plot = combined_fig_1abc,
  width = 14, height = 13, dpi = 300
)

ggsave(
  filename = here::here("output", "Figure_1.pdf"),
  plot = combined_fig_1abc,
  width = 14, height = 13, dpi = 1000
)

```




## 4.3. Fig. 2 - Stomatal response curves
```{r fig 2 curves of all species}

# Use posterior mean predictions generated in Section 2.2.
# These represent population-level curves without tree/leaf random effects.
curves_all <- posterior_means %>%
  left_join(
    data %>% select(species, tree_group) %>% distinct(),
    by = "species"
  ) %>%
# and add the new names!
     mutate(species = case_when(species=="Pla_ace" ~ "Pla_his", 
                            species=="Sor_dom" ~ "Cor_dom",
                            species=="Sor_tor" ~ "Tor_gla",
                            TRUE ~ species))

label_df <- curves_all %>%
  group_by(species) %>%
  slice_max(gsw_mean, n = 1) %>%
  ungroup()

# Classify species into low, intermediate, and high VPD optimum groups
# based on tertiles of the estimated VPD optimum.
quantiles <- quantile(stomatal_behaviour_species_metrics$VPD_opt, probs = c(1/3, 2/3), na.rm = TRUE)


species_groups_optimum <- stomatal_behaviour_species_metrics %>%
  mutate(optimum_group = case_when(
    VPD_opt >= quantiles[2] ~ "High VPD Optimum (VPD<sub>L</sub>>1)",
    VPD_opt < quantiles[2] & VPD_opt >= quantiles[1] ~ "Mid VPD Optimum (VPD<sub>L</sub>~1)",
    VPD_opt < quantiles[1] ~ "Low VPD Optimum (VPD<sub>L</sub><1)"
  )) %>%
  mutate(optimum_group = factor(optimum_group, levels = c(
    "Low VPD Optimum (VPD<sub>L</sub><1)",
    "Mid VPD Optimum (VPD<sub>L</sub>~1)",
    "High VPD Optimum (VPD<sub>L</sub>>1)"
  )))


# Add VPD optimum classification to the curve data for faceting
curves_vpd_grouped <- curves_all %>% 
  left_join(species_groups_optimum %>% 
              dplyr::select(species, optimum_group), by = "species")

# Generate species-specific colour gradients within each VPD optimum group.

group_counts <- species_groups_optimum %>% count(optimum_group)

n_high <- group_counts %>% filter(optimum_group == "High VPD Optimum (VPD<sub>L</sub>>1)") %>% pull(n)
n_mid  <- group_counts %>% filter(optimum_group == "Mid VPD Optimum (VPD<sub>L</sub>~1)")  %>% pull(n)
n_low  <- group_counts %>% filter(optimum_group == "Low VPD Optimum (VPD<sub>L</sub><1)")  %>% pull(n)

low_base  <- c("#EAB308", "#CA8A04", "#A16207", "#854D0E", "#713F12")
mid_base  <- c("#F97316", "#EA580C", "#C2410C", "#9A3412", "#7C2D12")
high_base <- c("#EF4444", "#DC2626", "#B91C1C", "#991B1B", "#450A0A")

colors_high <- colorRampPalette(high_base)(n_high)
colors_mid  <- colorRampPalette(mid_base)(n_mid)
colors_low  <- colorRampPalette(low_base)(n_low)

species_color_mapping <- species_groups_optimum %>%
  arrange(optimum_group, species) %>%
  group_by(optimum_group) %>%
  mutate(species_color = case_when(
    optimum_group == "High VPD Optimum (VPD<sub>L</sub>>1)" ~ colorRampPalette(high_base)(n()),
    optimum_group == "Mid VPD Optimum (VPD<sub>L</sub>~1)"  ~ colorRampPalette(mid_base)(n()),
    optimum_group == "Low VPD Optimum (VPD<sub>L</sub><1)"  ~ colorRampPalette(low_base)(n())
  )) %>%
  ungroup()

custom_colors <- species_color_mapping$species_color
names(custom_colors) <- species_color_mapping$species

# Determine curve peaks as label anchor points.
# Labels are subsequently displaced vertically to avoid overlap.
label_df_peak <- curves_vpd_grouped %>%
  group_by(species, optimum_group) %>%
  slice_max(gsw_mean, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(optimum_group, desc(gsw_mean)) %>%
  group_by(optimum_group) %>%
  mutate(
    x_label = VPD + 0.3,
    y_label = gsw_mean
  )

min_gap <- 0.02

label_df_peak <- label_df_peak %>%
  arrange(optimum_group, desc(y_label)) %>%
  group_by(optimum_group) %>%
  mutate(
    y_label = accumulate(
      y_label,
      ~ {
        if (.y > .x - min_gap) {
          .x - min_gap
        } else {
          .y
        }
      }
    )
  ) %>%
  mutate(
    displacement = abs(y_label - gsw_mean),
    x_label = VPD + 0.15 + 4 * displacement
  )

# Calculate the min and max VPD_opt for each facet panel
bg_rects <- species_groups_optimum %>%
  group_by(optimum_group) %>%
  summarise(
    xmin = min(VPD_opt, na.rm = TRUE),
    xmax = max(VPD_opt, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  # Inf and -Inf map to the bottom and top boundaries of the panel automatically
  mutate(
    ymin = -Inf,
    ymax = Inf
  )

# Plot population-level stomatal response curves grouped by VPD optimum.
# Curves show predicted stomatal conductance responses across the VPD gradient.
p_vpd_grouped <- 
  ggplot(curves_vpd_grouped, 
         aes(x = VPD, y = gsw_mean, colour = species)) +
  
  geom_rect(
    data = bg_rects,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "gray70",      
    alpha = 0.5,          
    inherit.aes = FALSE
  ) +
 
  geom_line(linewidth = 0.8, alpha = 0.85) +
  
  geom_segment(
    data = label_df_peak,
    aes(
    x = VPD,
    y = gsw_mean,
    xend = x_label,
    yend = y_label,
    colour = species),
    linewidth = 0.3
    ) +
  geom_label(
    data = label_df_peak,
    aes(
    x = x_label,
    y = y_label,
    label = species,
    colour = species),
    hjust = 0,
    size = 3.6,
    fill = alpha("white", 0.6),
    linewidth = NA,
    label.padding = unit(0.1, "lines")
    ) +
  
  facet_wrap(~optimum_group, ncol = 3) +
  
  scale_colour_manual(
    values = custom_colors, 
    guide = "none"
    ) +
  labs(
    x = expression("Vapor Pressure Deficit, VPD"["L"] * " (kPa)"),
    y = expression("Stomatal Conductance, g"["s"] * " (mol m"^-2 * " s"^-1 * ")")
    ) +
  coord_cartesian(clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme(
    plot.margin = margin(5.5, 60, 5.5, 5.5)
    ) +
  theme_bw(base_size=20) +
  theme(
    strip.background = element_blank(),
    strip.text = ggtext::element_markdown(face = "bold", size = 16, hjust = 0),
    panel.grid = element_blank()
  )

p_vpd_grouped

ggsave(
  here::here("output", "Fig_3_all_curves.png"),
  p_vpd_grouped,
  width = 11,
  height = 8.5, 
  dpi = 300
)


# save for submission
ggsave(
  here::here("output", "Figure_3.pdf"),
  p_vpd_grouped,
  width = 11,
  height = 8.5, 
  dpi = 1000
)

```


## 4.4. Fig. 3 - Stomatal Behaviour metrics

```{r figure 3 - all metrics in one space}

stomatal_behaviour_species_metrics <- 
  stomatal_behaviour_species_metrics %>%
  dplyr::select(-c(tree_group)) %>% 
  left_join(tree_group_column)

# Plot ranges and scaling

x_limits <- range(stomatal_behaviour_species_metrics$VPD_opt, na.rm = TRUE)
y_limits <- range(stomatal_behaviour_species_metrics$gs_max, na.rm = TRUE)

x_span <- diff(x_limits)
y_span <- diff(y_limits)

x_offset <- 0.05 * x_span
y_offset <- 0.05 * y_span


# plot geometry
plot_width <- 9
plot_height <- 6


# Create custom point symbols
#
# Each species is represented by a circle:
#   - outline colour = VPD50
#   - fill encodes stomatal closure rate (Sclosure):
#       shallow  -> empty circle
#       moderate -> half-filled circle
#       steep    -> filled circle
#
# Since ggplot does not provide half-filled points with independent outline
# and fill colours, the symbols are constructed manually from polygons.

circle_pts <- function(x0, y0, rx, ry, from = 0, to = 2*pi, n = 60) {
  th <- seq(from, to, length.out = n)
  data.frame(x = x0 + rx * cos(th), y = y0 + ry * sin(th))
}

aspect_correction <- 0.75   # tune up/down until circles look round

rx <- x_span * 0.009        # smaller base size (was 0.012)
ry <- rx * aspect_correction * (plot_width * y_span) / (x_span * plot_height)

# create 3 groups of the variable Sclosure
stomatal_behaviour_species_metrics <- 
  stomatal_behaviour_species_metrics %>% 
    mutate(
    Sclosure_group = ntile(Sclosure, 3),
    Sclosure_group = factor(
      Sclosure_group,
      levels = c(1, 2, 3),
      labels = c("shallow", "moderate", "steep")
    )
  )


# outline: always a full circle, for every point
point_outline <- stomatal_behaviour_species_metrics %>%
  filter(!is.na(Sclosure_group)) %>%
  mutate(point_id = row_number()) %>%
  group_by(point_id) %>%
  group_modify(~ {
    d <- .x
    coords <- circle_pts(d$VPD_opt, d$gs_max, rx, ry)
    coords$VPD50 <- d$VPD50
    coords
  }) %>%
  ungroup()




# fill: empty (shallow, skip), half-disc (moderate), full disc (steep)
point_fill <- stomatal_behaviour_species_metrics %>%
  filter(Sclosure_group %in% c("moderate", "steep")) %>%
  mutate(point_id = row_number()) %>%
  group_by(point_id) %>%
  group_modify(~ {
    d <- .x
    coords <- if (d$Sclosure_group == "steep") {
      circle_pts(d$VPD_opt, d$gs_max, rx, ry)               # full disc
    } else {
rbind(
  circle_pts(d$VPD_opt, d$gs_max, rx, ry, from = pi, to = 2*pi),  # now lower half
  data.frame(x = d$VPD_opt, y = d$gs_max)
)
    }
    coords$VPD50 <- d$VPD50
    coords
  }) %>%
  ungroup()




vpd50_range <- range(stomatal_behaviour_species_metrics$VPD50, na.rm = TRUE)

# Midpoints of the plotting range used to divide the trait space
# into four conceptual quadrants.
mid_VPDopt <- mean(range(stomatal_behaviour_species_metrics$VPD_opt, na.rm = TRUE))
mid_gs <- mean(range(stomatal_behaviour_species_metrics$gs_max, na.rm = TRUE))


# Positions and alignment of the four quadrant labels.
# The labels are positioned slightly outside the data range so that
# left labels start at the left edge and right labels end at the right edge.
quad_labels <- tibble::tibble(
  label = c(
    "High-capacity\nearly-peaking",
    "High-capacity\nlate-peaking",
    "Low-capacity\nearly-peaking",
    "Low-capacity\nlate-peaking"
  ),
  
x = c(
  x_limits[1] - x_offset,
  x_limits[2] + x_offset,
  x_limits[1] - x_offset,
  x_limits[2] + x_offset
),
  
y = c(
  y_limits[2] + y_offset,
  y_limits[2] + y_offset,
  y_limits[1] - y_offset,
  y_limits[1] - y_offset
),
  
  hjust = c(0, 1, 0, 1),
  vjust = c(1, 1, 0, 0)
)

# create the figure
fig_all_metrics <- stomatal_behaviour_species_metrics %>%
  ggplot(aes(VPD_opt, gs_max)) +
  geom_vline(xintercept = mid_VPDopt, colour = "grey60", linewidth = 0.4, linetype = "dotted") +
  geom_hline(yintercept = mid_gs,  colour = "grey60", linewidth = 0.4, linetype = "dotted") +
  geom_text(data = quad_labels,
            aes(x=x, y=y, label = label, hjust = hjust, vjust = vjust),
            size = 5, fontface = "italic", colour = "grey35", lineheight = 0.9) +

  geom_polygon(data = point_fill, aes(x, y, group = point_id, fill = VPD50)) +
  geom_path(data = point_outline, aes(x, y, group = point_id, colour = VPD50),
            linewidth = 1) +

scale_fill_viridis_c(
  option = "magma", direction = -1, end = 0.9,
  limits = vpd50_range,
  guide = "none"
) +
scale_colour_viridis_c(
  name   = expression(atop("VPD at half-closure:", "" * VPD[50] * " (kPa)")),
  option = "magma", direction = -1, end = 0.9,
  limits = vpd50_range,
  guide  = guide_colorbar(order = 1, title.position = "top", title.hjust = 0.5)
) +
geom_point(data = stomatal_behaviour_species_metrics,
           aes(x = VPD_opt, y = gs_max, shape = Sclosure_group),
           colour = "black", fill = "black", size = 4,
           alpha = 0, inherit.aes = FALSE) +

scale_shape_manual(
  name   = expression(atop("Closure rate:", "" * S[closure] * "")),
  values = c("shallow" = 21, "moderate" = 21, "steep" = 21),
  guide  = guide_legend(order = 2, title.position = "top", title.hjust = 0.5,
                         override.aes = list(size = 3, alpha = 0))
) +
  new_scale_colour() +
  
  geom_text_repel(aes(
                  label = species, 
                  colour = tree_group, 
                  segment.colour=tree_group),
                  fontface="italic",
                  size = 3.5,
                  max.overlaps = Inf,
                  force = 2, force_pull = 0.5,
                  nudge_x = 0.01, nudge_y = 0.01,
                  box.padding = 0.3, point.padding = 0.1,
                  min.segment.length = 0.3,
                  segment.size = 0.3, seed = 42) +
  scale_colour_manual(
    name = "Tree group",
    aesthetics = c("colour", "segment.colour"),
    values = c("broadleaved" = "black", "conifer" = "grey50"),
    guide= "none"
    # guide = guide_legend(order = 3, title.position = "top", title.hjust = 0.5)
  ) +

  scale_x_continuous(
    name = expression("VPD optimum (VPD"[opt]*", kPa)")) +
  scale_y_continuous(
    name = expression("Maximum "*italic(g)[s]*" ("*italic(g)[max]*", mol m"^{-2}*" s"^{-1}*")")) +
  
  theme_classic(base_size = 11) +
  theme(
    legend.position = "right",
    legend.key.size = unit(0.5, "cm"),
    legend.title    = element_text(size = 14, hjust = 0.5),
    legend.text     = element_text(size = 12),
    axis.title      = element_text(size=16),
    axis.line       = element_line(colour = "grey40"),
    panel.border    = element_rect(colour = "grey40", fill = NA),
    plot.margin     = margin(20, 20, 10, 10)
  )

ggsave(here("output", "Fig_all_metrics.png"), fig_all_metrics, height=6, width=9.5, dpi=300)

ggsave(here("output", "Fig_all_metrics.svg"), fig_all_metrics, height=6, width=9.5, dpi=1000)

```




# 5. Figures (Supplementary)
## 5.1. Fig. S6 - Prior predictive check

```{r prior check}

baseline <- create_prediction_grid(
  sp = species_list[1],
  VPD_grid = VPD_grid_plot
)

# Extract predictor draws
prior_samples_grid <- posterior_epred(
  model_check_priors,
  newdata = baseline,
  transform = TRUE,
  re_formula = NA  
)

# Sample 100 draws and format for ggplot
set.seed(123)
draw_indices <- sample(1:nrow(prior_samples_grid), 100)
pp_curves_sampled <- prior_samples_grid[draw_indices, ]

# Convert the matrix to a long-format data frame for ggplot
plot_data <- as.data.frame(t(pp_curves_sampled)) %>%
  mutate(VPDleaf = VPD_grid_plot) %>%
  pivot_longer(
    cols = -VPDleaf, 
    names_to = "draw", 
    values_to = "gsw"
  )

fig_prior_predictive <- ggplot(plot_data, aes(x = VPDleaf, y = gsw, group = draw)) +
  geom_line(alpha = 0.3, color = "royalblue", linewidth = 0.5) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(expand = c(0, 0)) + 
  labs(
    x = expression(Vapor~Pressure~Deficit~(VPD[L]*", kPa")), # Added typical units
    y = expression(Stomatal~Conductance~(italic(g)[s]*", mol "~m^{-2}~s^{-1}))
  ) +
  theme_minimal(base_size = 11) + 
  theme(
    panel.grid.minor = element_blank(),
    axis.title = element_text(face = "plain")
  )

ggsave(here("output", "Fig_S6_Priors.png"), fig_prior_predictive, dpi=300, height=5, width=7)

```


## 5.2. Fig. S5 - Curves of single species
```{r single species curve}

# Create the single faceted plot
facet_plot <- ggplot() +
  
  geom_point(data = data, aes(VPDleaf, gsw, colour=as.factor(tree_individual)), alpha = 0.4, size = 0.6) +
  
  
  geom_line(data = posterior_draws, aes(VPD, gsw, group = draw), alpha = 0.15, linewidth = 0.3, color = "grey70") +
  geom_line(data = posterior_means, aes(VPD, gsw_mean), linewidth = 1.0, color = "black") +
  
  facet_wrap(~species, ncol = 5, scales = "free_y") +
  scale_color_manual(values = pastel) +
  labs(
    x = expression("VPD"["L"] * " (kPa)"),
    y = expression("g"["s"] * " (mol m"^-2 * " s"^-1 * ")")
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "italic", size = 9),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 8)
  )

ggsave(
  filename = here::here("output", "Fig_S5_species_curves.png"),
  plot = facet_plot,
  width = 6.5,
  height = 9,
  dpi = 300
)

```




##5.3. Fig. S8 - Gymno vs. Angio

```{r fig metrics gymno vs. angio}

fig_metric_diff_tree_group

ggsave(
  filename = here::here(
    "output",
    "Fig_SX_clade_differences_in_metrics.png"
  ),
  plot = fig_metric_diff_tree_group,
  width = 9,
  height = 5,
  dpi = 300
)

```

## 5.4. Fig. S9 - Metrics correlation

```{r correlation matrix}

correlation_metrics_data <- stomatal_behaviour_species_metrics %>%
  dplyr::select(gs_max, VPD_opt, VPD50, Sclosure)

res <- Hmisc::rcorr(
  as.matrix(correlation_metrics_data),
  type = "pearson"
)

cor_long <- as.data.frame(as.table(res$r)) %>%
  rename(Var1 = Var1, Var2 = Var2, correlation = Freq)

p_long <- as.data.frame(as.table(res$P)) %>%
  rename(Var1 = Var1, Var2 = Var2, p_value = Freq)

plot_data <- left_join(cor_long, p_long, by = c("Var1", "Var2")) %>%
  mutate(
    sig_label = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      p_value < 0.1   ~ ".",
      TRUE ~ ""
    )
  )

diagonal_data <- plot_data %>% filter(Var1 == Var2)

var_labels <- c(
  gs_max   = expression(g[max]),
  VPD_opt  = expression(VPD[opt]),
  VPD50    = expression(VPD[50]),
  Sclosure = expression(S[closure])
)

fig_corr_metrics <- ggplot(plot_data, aes(x = Var2, y = Var1, fill = correlation)) +
  geom_tile(color = "white") +
  geom_tile(data = diagonal_data, fill = "grey40", color = "white") +  # grey diagonal
  geom_text(aes(label = paste0(round(correlation, 2))),
            color = "white", size = 4, lineheight = 0.9) +
  scale_fill_gradient2(
    low = "#d6604d", mid = "grey90", high = "#2166ac",
    midpoint = 0, limits = c(-1, 1),
    name = "Pearson r"
  ) +
  scale_x_discrete(labels = var_labels) +
  scale_y_discrete(labels = var_labels) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    legend.position = "right"
  ) +
  coord_fixed()

fig_corr_metrics

ggsave(here("output", "Fig_S8_correlation_metrics.png"), fig_corr_metrics, height=6, width=9)

```

## 5.5. Fig. S10 - Coordination of metrics
```{r coordination of metrics}

# Prepare species-level metric matrix for PCA
# Species become row names because prcomp() expects observations in rows.
data_pca <- stomatal_behaviour_species_metrics %>%
  dplyr::select(
    species,
    gs_max, 
    VPD_opt,
    VPD50, 
    Sclosure
  ) %>%
  column_to_rownames("species")



# Run PCA on standardized variables
# Scaling ensures that all metrics contribute equally despite different units.
pca_res <- prcomp(data_pca, center = TRUE, scale. = TRUE)

# Extract PCA scores (species positions) for plotting
ind_coords <- as.data.frame(pca_res$x) %>%
  rownames_to_column("species") %>%
  left_join(tree_group_column, by = "species")

# Extract PCA loadings (metric vectors)
var_coords <- as.data.frame(pca_res$rotation)
var_coords$var_names <- rownames(var_coords)

# Vector length in the PC1-PC2 plane.
# Used to scale arrow linewidths if desired.
var_coords$contrib <- sqrt(var_coords$PC1^2 + var_coords$PC2^2)

# Scale loading vectors to fit into the same coordinate system
# as the species scores for visualization.
scale_factor <- min(
  diff(range(ind_coords$PC1)) / diff(range(var_coords$PC1)),
  diff(range(ind_coords$PC2)) / diff(range(var_coords$PC2))
) * 0.7

var_coords$PC1_s <- var_coords$PC1 * scale_factor
var_coords$PC2_s <- var_coords$PC2 * scale_factor


# Position labels slightly beyond the arrow tips
# so they do not overlap with the arrowheads.
var_coords$angle <- atan2(var_coords$PC2_s, var_coords$PC1_s)
offset <- 0.06 * max(abs(c(var_coords$PC1_s, var_coords$PC2_s)))

var_coords$x_lab <- var_coords$PC1_s + offset * cos(var_coords$angle)
var_coords$y_lab <- var_coords$PC2_s + offset * sin(var_coords$angle)

# Expressions parsed by ggplot to format subscripts and italics.
var_coords$var_label <- c(
  gs_max   = "bolditalic(g)[s*' max']",
  VPD_opt  = "bold(VPD)[opt]",
  VPD50    = "bold(VPD)[50]",
  Sclosure = "bold(S)[closure]"
)[var_coords$var_names]


# Percentage of variance explained by each principal component
# for axis labels.
pca_var <- (pca_res$sdev^2) / sum(pca_res$sdev^2) * 100

pc1_lab <- paste0("PC1 (", round(pca_var[1], 1), "%)")
pc2_lab <- paste0("PC2 (", round(pca_var[2], 1), "%)")

# PCA biplot 
pca_metrics <- 
  
  ggplot(ind_coords, aes(PC1, PC2)) +

  geom_point(
    aes(shape = tree_group, colour=tree_group),
    size = 3.5,
    alpha = 0.9
    ) +
  geom_segment(
  data = var_coords,
  aes(
    x = 0, y = 0,
    xend = PC1_s,
    yend = PC2_s
  ),
  arrow = arrow(length = unit(0.22, "cm")),
  color = "black",
  alpha = 0.9
  ) +
  geom_label_repel(
    data = var_coords,
    aes(x = PC1_s, y = PC2_s, label = var_label),
    parse = TRUE,

  fill = scales::alpha("white", 0.7),

  label.size = NA,      # <- removes border completely
  label.padding = unit(0.15, "lines"),

  box.padding = 0.3,
  point.padding = 0.2,
  segment.color = NA,

  size = 3.5,
  fontface = "bold"
  ) +

  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.4) +

  coord_fixed(ratio = 1) +
  
    scale_color_manual(
      values = c("broadleaved" = "#41624D", "conifer" = "#7798B9"),
      name= "Tree group") +

  scale_shape_manual(
    values = c(
      "broadleaved" = 16,
      "conifer" = 17
    ),
    name = "Tree group"
  ) +

  labs(
    x = pc1_lab,
    y = pc2_lab,
  ) +

  theme_minimal(base_size = 13) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right"
  )

pca_metrics

ggsave(
  filename = here::here("output", "Fig_S10_Coordination_metrics.png"),
  plot = pca_metrics,
  width = 7,
  height = 4,
  dpi = 300
)



```
## 5.6. Fig. S7 - Overview of Metrics

```{r metrics figure}

# stomatal_behaviour_species_metrics <- 
#   stomatal_behaviour_species_metrics %>% 
#   mutate(tree_group = case_when(
#     species=="Tor_gla" ~ "broadleaved",
#     species=="Pla_his" ~ "broadleaved",
#     species=="Cor_dom" ~ "broadleaved",
#     TRUE ~ tree_group))


species_order <- stomatal_behaviour_species_metrics %>%
  arrange(gs_max) %>% 
  pull(species)

metrics_long <- stomatal_behaviour_species_metrics %>%
  rename(
    gs_mean       = gs_max,
    VPD_mean      = VPD_opt,
    VPD50_mean    = VPD50,
    Sclosure_mean = Sclosure
  ) %>%
  dplyr::select(
    species, 
    tree_group, 
    gs_mean, gs_l95 = gs_max_l95, gs_u95 = gs_max_u95,
    VPD_mean, VPD_l95 = VPD_opt_l95, VPD_u95 = VPD_opt_u95,
    VPD50_mean, VPD50_l95, VPD50_u95,
    Sclosure_mean, Sclosure_l95, Sclosure_u95
  ) %>%
  pivot_longer(
    cols = -c(species, tree_group), 
    names_to = c("metric", ".value"),
    names_sep = "_"
  ) %>%
  mutate(
    metric = factor(metric, 
                    levels = c("gs", "VPD", "VPD50", "Sclosure"),
                    labels = c("gs_max", "VPD_opt", "VPD_50", "S_closure")),
    # FIX: Force text to lowercase so it strictly matches scale_shape_manual keys
    tree_group = tolower(trimws(tree_group)) 
  ) %>%
  arrange(metric, mean) %>% 
  distinct() %>% 
  ungroup() %>%
  mutate(
    species = factor(species, levels = species_order),
    tree_group = stringr::str_to_title(tree_group) 
  ) %>% 
  group_by(metric) %>%
  # Calculate the across-species mean for each metric
  # (displayed as the vertical reference line).
  mutate(panel_mean = mean(mean, na.rm = TRUE)) %>%
  ungroup()


metrics_list <- split(metrics_long, metrics_long$metric)

make_panel <- function(data, title_text, x_axis_expr, y_title = NULL) {
  
  ggplot(data, aes(x = mean, y = fct_reorder(species, mean))) +
    geom_vline(
      aes(xintercept = panel_mean),
      colour = "black",
      linewidth = 0.4,
      linetype = "dotted"
    ) +
    geom_errorbar(aes(xmin = l95, xmax = u95), 
                  color = "gray40", width = 0, linewidth = 0.4) +
    geom_point(aes(shape = tree_group, color=tree_group), 
               size = 2.5, stroke = 0.5) + 
    scale_shape_manual(
      values = c("Broadleaved" = 16, "Conifer" = 17),
      name = ""
    ) +  
    scale_color_manual(
      values = c("Broadleaved" = "#41624D", "Conifer" = "#7798B9")
    ) +
    guides(
      colour = "none",
      shape = guide_legend(
        override.aes = list(
          colour = c("#41624D", "#7798B9")
        )
      )
    ) +
    
    labs(title = title_text, x = x_axis_expr, y = y_title) +
    theme_bw(base_size = 14) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(color = "gray96"),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14), 
      axis.text.y = element_text(face = "italic"),
      
      legend.position = "bottom"
    )
}


# Create one panel for each stomatal behaviour metric.
p1 <- make_panel(metrics_list$gs_max,    "Maximum gs", expression(g[max]~(mol~m^{-2}~s^{-1})), "Species")
p2 <- make_panel(metrics_list$VPD_opt,   "VPD optimum",      expression(VPD[opt]~(kPa)))
p3 <- make_panel(metrics_list$VPD_50,    "VPD at half-closure",    expression(VPD[50]~(kPa)))
p4 <- make_panel(metrics_list$S_closure, "Closure rate",     expression(S[closure]))


fig_metrics <- (p1 | p2 | p3 | p4) +
  plot_layout(guides = "collect") +
  plot_annotation(
    theme = theme(
      plot.margin = margin(10, 10, 10, 10),
      legend.position = "bottom",
      legend.text = element_text(size = 20),
      legend.title = element_text(size = 20),
      legend.key.size = unit(2, "lines")
    )
  )


ggsave(
  filename = here::here("output", "Fig_S7_metrics.png"),
  plot = fig_metrics,
  width = 9.5,
  height = 8,
  dpi = 300
)

```



