---
title: "Stomatal Behaviour - 03 Trait Analysis"
author: "Lena Sachsenmaier"
output:
  html_document:
    df_print: paged
contact: lena.sachsenmaier.research@gmail.de
created: "2026-07-30"
updated: "`r format(Sys.time(), '%d %B, %Y')`"
---


# Analysis of stomatal behaviour metrics and their relationships with
# leaf traits and drought response responses

```{r setup, include=FALSE}

library(tidyverse) # basic data handling
library(here) # reproducible file pathes

# data viualization and figure arrangement
library(ggExtra)
library(ggpubr)
library(patchwork)
library(ggeffects)
library(ggrepel)
library(scales)  # for rescale()
library(ggdist)

# modeling, model comparison and effect viualization
library(brms)
library(loo)
library(tidybayes)
library(bayestestR)
library(emmeans) # estimated marginal means and contrasts
library(MuMIn) # model selection and R2 calculations

# Correlation, PCA and PLS analysis
library(corrplot)
library(factoextra)
library(pls)

# exporting results tables
library(writexl)

# always display the output at the end of the chunk
knitr::opts_chunk$set(
  results = "hold")

```

# 1.Preparation of data frames

## 1.1 Load the data frames

* stomatal control metrics
* trait set: stomata density & size
* trait set: leaf economic spectrum traits (LES)
* drought response: leaf Dd13C 2018 - 2021
* list of trees measured during the stomatal control campaign

```{r load data, results="hide"}

# We select only the trait measurements and leaf carbon isotope measurements of the very same trees

# load the cleaned stomatal conductance data again, because we need
# the IDs of the trees that have been measured during the stomatal conductance campaign
trees_measured <- read_csv(here("data", "processed", "data_cleaned.csv")) %>% 
  dplyr::select(tree_group, species, tree_id, tree_group) %>% 
  distinct()

tree_group_column <- trees_measured %>% dplyr::select(c(tree_group, species)) %>% distinct()


# (1) load the stomatal behaviour metrics data with all draws and only species means:
stomatal_behaviour_metrics <- 
  read_csv(here("data", "processed", "stomatal_behaviour_metrics.csv"))

stomatal_behaviour_species_metrics <- 
  read_csv(here("data", "processed", "stomatal_behaviour_species_metrics.csv"))


# (2) stomata anatomy data
# angiosperm stomata
traits_stomata_angio <- 
  read_csv(here("data", "raw", "Kretz_Sachsenmaier_et_al_2026_angiosperm_stomatal_trait_data.csv")) %>% 
  unite(tree_id, species_code, block, sep="-", remove=FALSE) %>% 
    rename(species_full = species,
         species = species_code) %>% 
  dplyr::select(-c(GCPW, PW)) %>% 
  # just for verifying that in the dataset of the stomata trait data are only trees for which gs was measured
  semi_join(trees_measured, by = "tree_id")

# gymnosperm stomata
traits_stomata_gymno <- 
  read_csv(here("data", "raw", "Sachsenmaier_et_al_2026_conifer_stomatal_trait_data.csv")) %>% 
  dplyr::select(-c(SCL, SCW, SD_ab, SD_ad, SD_sbs, image_no_100x, image_no_1000x)) %>% 
  # For gymnosperms, LI-600N measurements integrate all needle surfaces (because of the differnt chamber);
  # therefore full-surface stomatal density (SD_fs) is used.
  rename(SD = SD_fs) %>% 
    separate(species, into = c("genus", "epithet"), sep = " ", remove = FALSE) %>%
  mutate(species_code = paste0(substr(genus, 1, 3), "_", substr(epithet, 1, 3))) %>%
  relocate(species_code, .after = species) %>% 
  rename(species_full = species,
         species = species_code) %>% 
  dplyr::select(-c(epithet)) %>% 
  # kick out all the tree individuals for which gs was NOT measured
  mutate(kick_out = case_when(
    tree_id== "Abi_alb_25_13" |
    tree_id == "Lar_dec_25_12" |
    tree_id == "Pin_nig_21_9" |
    tree_id=="Pse_men_28_12" ~ TRUE,
    species == "Tsu_can" ~ TRUE,
    TRUE ~ FALSE)) %>% 
  filter(kick_out==FALSE) %>% 
  dplyr::select(-c(kick_out, GCPW, PW, genus))
  
  
# (3) load the leaf functional trait data set:
traits_leaf_morph <- 
  read_csv(here("data", "raw","data_leaf_traits.csv")) %>% 
  dplyr::select(c(species, block_ID, SLA, LDMC, 
                  leaf_N_area, thickness, toughness)) %>%
  separate(species, into = c("genus", "epithet"), sep = " ", remove = FALSE) %>%
  mutate(sp = paste0(substr(genus, 1, 3), "_", substr(epithet, 1, 3))) %>% 
  dplyr::select(-c(epithet)) %>% 
    rename(species_full = species,
           species = sp) %>% 
  relocate(species, .after=species_full) %>% 
  unite(tree_id, species, block_ID, sep="-", remove=FALSE) %>% 
  # only keep the trees for which we also measured gs
  semi_join(trees_measured, by = "tree_id")
  

# (4) leaf carbon isotope data set
delta_d13C_species_means <- read_csv(here("data", "raw", "data_leaf_carbon_isotopes_2018_2021.csv"))


```

## 1.2. Preparation of the data frames

Only retain trait measurements from individuals included in the
stomatal conductance campaign to ensure trait and physiological
measurements refer to the same sampled trees.

### 1.2.1. Trait data frame

```{r traits data frame}

# join the stomata traits of gymno- and angiosperms
# and calculate the species means
traits_stomata <- traits_stomata_angio %>%
  bind_rows(traits_stomata_gymno)

rm(traits_stomata_angio, traits_stomata_gymno)


traits_stomata_sp <- traits_stomata %>% 
  group_by(species, species_full) %>% 
  dplyr::summarize(
            SD_sd = sd(SD),
            GCL_sd = sd(GCL, na.rm=TRUE),
            SD = mean(SD),
            GCL = mean(GCL, na.rm=TRUE)
) %>% 
  ungroup() 


# Leaf Economic Traits

# Restrict the dataset to the trees used in the stomatal conductance campaign.
# This ensures that leaf traits and stomatal behaviour metrics originate from the same
# individuals before calculating species-level means.

traits_leaf_morph_sp <- 
  traits_leaf_morph %>%
  group_by(species) %>% 
  dplyr::summarize(
            SLA_sd = sd(SLA),
            LDMC_sd = sd(LDMC),
            SLA = mean(SLA),
            LDMC = mean(LDMC),
            leaf_N_area_sd = sd(leaf_N_area),
            leaf_N_area = mean(leaf_N_area),
            thick_sd = sd(thickness),
            thick = mean(thickness),
            tough_sd = sd(toughness),
            tough = mean(toughness)
  ) %>% 
  ungroup()



# one full traits data frame:

traits_df <- traits_stomata_sp %>% 
  left_join(traits_leaf_morph_sp)


```


### 1.2.2. Physiological Drought Response 

```{r Delta d13C df}

drought_response_sp <- delta_d13C_species_means %>% dplyr::select(c(species, delta_d13C))

```


### 1.2.3. Data frames for models
```{r dataframe for models}

# Combine stomatal behaviour metrics, leaf traits, and drought response data.
# Stomatal metrics contain posterior draws, whereas traits and δ13C are
# species-level measurements.
data_draws_full <- stomatal_behaviour_metrics %>%  
  left_join(traits_df) %>% 
  left_join(drought_response_sp) %>% 
  left_join(tree_group_column)

set.seed(123)
# Randomly subsample posterior draws per species for downstream plotting
# and trait regressions. This retains uncertainty while reducing object size.
data_draws_sampled <- data_draws_full %>%
  group_by(species) %>%
  slice_sample(n = 100) %>%
  ungroup()


# only the mean dataframe
data_mean <- 
  stomatal_behaviour_species_metrics %>%
    left_join(tree_group_column) %>% 
    left_join(traits_df) %>% 
    left_join(drought_response_sp)


# Aggregate posterior draws to species level while retaining uncertainty.
# The standard deviation of stomatal metrics represents posterior uncertainty,
# which is later propagated in Bayesian models using se().

data_for_models <- 
  data_draws_full %>%
  group_by(tree_group, species) %>%
  summarise(
    gs_max_mean = mean(gs_max),
    gs_max_sd   = sd(gs_max),
    
    VPD_opt_mean = mean(VPD_opt),
    VPD_opt_sd   = sd(VPD_opt),
    
    VPD50_mean = mean(VPD50),
    VPD50_sd   = sd(VPD50),
    
    Sclosure_mean = mean(Sclosure),
    Sclosure_sd   = sd(Sclosure),
    
    # Leaf traits
    SD = mean(SD),
    GCL = mean(GCL),
    SLA = mean(SLA),
    LDMC = mean(LDMC),
    leaf_N_area = mean(leaf_N_area),
    thick = mean(thick),
    tough = mean(tough),
    
    # Physiological drought response
    delta_d13C = mean(delta_d13C)
  ) %>% 
  ungroup() %>%
  mutate(
    tree_group = factor(tree_group),
    
    # Standardized predictors
    SD_scaled = as.numeric(scale(SD)),
    GCL_scaled = as.numeric(scale(GCL)),
    SLA_scaled = as.numeric(scale(SLA)),
    LDMC_scaled = as.numeric(scale(LDMC)),
    thick_scaled = as.numeric(scale(thick)),
    tough_scaled = as.numeric(scale(tough)),
    leaf_N_area_scaled = as.numeric(scale(leaf_N_area)),
    delta_d13C_scaled = as.numeric(scale(delta_d13C))
  )


```



### 1.2.4. Tab. S7 - Trait table
Nice table for Trait data (Supplementary Table S7)

```{r supplementary data tables}

# formatting function for the numbers in the supplementary table
format_num <- function(x, digits) {
  formatC(x, format = "f", digits = digits)
}

# Create nice table as supplementary

trait_table <- data_mean %>%
  dplyr::select(
    species,
    SD, SD_sd,
    GCL, GCL_sd,
    SLA, SLA_sd,
    LDMC, LDMC_sd,
    thick, thick_sd,
    tough, tough_sd,
    leaf_N_area, leaf_N_area_sd
  ) %>%
 mutate(
    SD = paste0(format_num(SD, 2), " (±", format_num(SD_sd, 2), ")"),
    GCL = paste0(format_num(GCL, 2), " (±", format_num(GCL_sd, 2), ")"),
    SLA = paste0(format_num(SLA, 2), " (±", format_num(SLA_sd, 2), ")"),
    LDMC = paste0(format_num(LDMC, 3), " (±", format_num(LDMC_sd, 3), ")"),
    thick = paste0(format_num(thick, 3), " (±", format_num(thick_sd, 3), ")"),
    tough = paste0(format_num(tough, 2), " (±", format_num(tough_sd, 2), ")"),
    leaf_N_area = paste0(format_num(leaf_N_area, 4), " (±", format_num(leaf_N_area_sd, 4), ")")
  ) %>%
  dplyr::select(
    species,
    SD,
    GCL,
    SLA,
    LDMC,
    thick,
    tough,
    leaf_N_area
  )

write_xlsx(
  trait_table,
  here("output","Tab_S7_Trait_table.xlsx"
))
```



# 2. Models

## 2.1. Function for models

```{r function for models}

run_metric_models <- function(metrics, predictors, df, suffix = "") {
  
  for (m in metrics) {
    for (p in predictors) {
    # Response variables consist of the posterior mean and uncertainty
    # of each stomatal behaviour metric.
      metric_mean_col <- paste0(m, "_mean")
      metric_sd_col <- paste0(m, "_sd")
      
      # Model names are used for saving fitted brms objects.
      model_name <- paste0("model_", m, "_", p, suffix)
          
      message("Running model for Metric: ", m, " and Predictor: ", p, "...")
    
    # Model structure:
    # stomatal metric uncertainty | trait × tree functional group interaction
    #
    # The interaction tests whether trait relationships differ between
    # broadleaved and conifer species.
    # sigma = TRUE means the provided uncertainty is known measurement error, but residual variance is still     # estimated.
      
      model_formula <- as.formula(
        paste0(
          metric_mean_col,
          " | se(", metric_sd_col, ", sigma = TRUE) ~ ",
          p,
          " * tree_group"
        )
      )
      # fit the model
      fit <- brm(
        formula = model_formula,
        data = df,
        chains = 4,
        iter = 4000, # changed to 4000 instead of 2000 (on 07.09.2026)
        cores = 4,
        file = here::here("output", "models", model_name),
        file_refit = "on_change"
      )
    # Store models with informative names for later extraction of results
    # and visualization.
      assign(model_name, fit, envir = .GlobalEnv)
    }
  }
}

```



## 2.2. Relationships between stomatal behaviour and leaf traits 

```{r trait models}
# Test whether leaf functional traits explain variation in stomatal behaviour metrics.
# For each combination of stomatal metric and trait, a Bayesian linear model is fitted.
# Species-level means are used as responses, while uncertainty in the estimated
# stomatal metrics is propagated using the standard deviation of posterior draws.

run_metric_models(
  metrics = c("gs_max", "VPD_opt", "VPD50", "Sclosure"),
  predictors = c("SD_scaled", "GCL_scaled", "SLA_scaled",
                 "LDMC_scaled", "thick_scaled",
                 "tough_scaled", "leaf_N_area_scaled"),
  df = data_for_models
)


```


## 2.3. Relationship between stomatal behaviour and leaf ∆δ13C

```{r d13C models}

# Test whether physiological drought response (δ13C) explains variation
# in stomatal behaviour metrics.
#
# Each model uses species-level stomatal metrics with propagated posterior
# uncertainty and tests whether the relationship differs between conifers
# and broadleaved species.

run_metric_models(
  metrics = c("gs_max", "VPD_opt", "VPD50", "Sclosure"),
  predictors = "delta_d13C_scaled",
  df = data_for_models
)

```




# 3. Main Figures
## 3.1. Fig. 6 - Leaf traits effect size figure
```{r leaf trait effect size figure}

metrics <- c("gs_max", "VPD_opt", "VPD50", "Sclosure")
traits  <- c("SD_scaled", "GCL_scaled", "SLA_scaled", "LDMC_scaled", 
             "thick_scaled", "tough_scaled", "leaf_N_area_scaled")

summary_list_trait_models <- list()
model_counter <- 1

draws_list_trait_models <- list()

for (m in metrics) {
  for (t in traits) {
    
    original_model_name <- paste0("model_", m, "_", t)
    file_path <- here::here("output", "models", paste0(original_model_name, ".rds"))
    
    if (!file.exists(file_path)) next
    
    fit <- readRDS(file_path)
    
    trait_scale_col <- t
    
    # Calculate Bayes R2
    r2_val <- bayes_R2(fit)[1, "Estimate"]
    
    # Clean up names
    clean_metric <- gsub("_scaled", "", m)
    clean_trait  <- gsub("_scaled", "", t)
    display_model_name <- paste0("Model ", model_counter)
    
    # Extract slopes via emtrends — same as your single panel figures
    trends <- tryCatch(
      as.data.frame(emtrends(fit, ~ tree_group, var = trait_scale_col)),
      error = function(e) { warning(paste("emtrends failed for", original_model_name)); return(NULL) }
    )
    
    if (is.null(trends)) next
    
    colnames(trends)[colnames(trends) == paste0(trait_scale_col, ".trend")] <- "trend"
    
 for (grp in c("broadleaved", "conifer")) {
  
  # Get full posterior draws for this group's slope
  slope_draws <- fit %>% 
    emtrends(~ tree_group, var = trait_scale_col) %>%
    gather_emmeans_draws() %>%
    filter(tree_group == grp) %>%
    rename(slope = .value)
  
  if (nrow(slope_draws) == 0) next
  
  draws_list_trait_models[[length(draws_list_trait_models) + 1]] <- slope_draws %>%
  mutate(
    model_name = display_model_name,
    metric     = clean_metric,
    trait      = clean_trait,
    tree_group = grp
  )
  
  # Summarise with 95% credible interval
  grp_summary <- slope_draws %>%
    tidybayes::median_qi(slope, .width = 0.95)
  
  summary_list_trait_models[[length(summary_list_trait_models) + 1]] <- data.frame(
    model_name = display_model_name,
    metric     = clean_metric,
    trait      = clean_trait,
    tree_group = grp,
    R2         = r2_val,
    slope      = grp_summary$slope,
    CI_lower   = grp_summary$.lower,
    CI_upper   = grp_summary$.upper,
    stringsAsFactors = FALSE
  )
}
    
    model_counter <- model_counter + 1
  }
}



# Combine and apply rounding rules
 
summary_table_trait_models <- 
  bind_rows(summary_list_trait_models) %>% 
  mutate(
    R2       = round(R2, 2),
    slope    = round(slope, 3),
    CI_lower = round(CI_lower, 2),
    CI_upper = round(CI_upper, 2)
  )


# Named vector for nice trait labels
trait_labels <- c(
  "SD"          = "Stomatal density",
  "GCL"         = "Guard cell length",
  "SLA"         = "Specific leaf area",
  "LDMC"        = "Leaf dry matter content",
  "thick"       = "Leaf thickness",
  "tough"       = "Leaf toughness",
  "leaf_N_area" = "Leaf nitrogen (per area)"
)

# Custom labeller for metrics using plotmath expressions
metric_labeller <- as_labeller(
  c(
    "gs_max"  = "g[max]",
    "VPD_opt"  = "VPD[opt]",
    "VPD50"     = "VPD[50]",
    "Sclosure"  = "S[closure]"
  ),
  default = label_parsed   # tells ggplot to parse as plotmath
)



correct_trait_levels  <- c("SD", "GCL", "SLA", "LDMC", "thick", "tough", "leaf_N_area")
correct_metric_levels <- c("gs_max", "VPD_opt", "VPD50", "Sclosure")

plot_data <- summary_table_trait_models %>%
  mutate(
    tree_group = factor(tree_group, levels = c("conifer", "broadleaved")),
    trait      = factor(trait,  levels = correct_trait_levels),
    metric     = factor(metric, levels = correct_metric_levels)
  )

all_draws <- bind_rows(draws_list_trait_models) %>%
  mutate(
    tree_group = factor(tree_group, levels = c("conifer", "broadleaved")),
    trait      = factor(trait,  levels = correct_trait_levels),
    metric     = factor(metric, levels = correct_metric_levels)
  )

highlight_panels <- summary_table_trait_models %>%
  group_by(metric, trait) %>%
  summarise(highlight = any(CI_lower > 0 | CI_upper < 0), .groups = "drop") %>%
  filter(highlight) %>%
  mutate(
    trait  = factor(trait,  levels = correct_trait_levels),
    metric = factor(metric, levels = correct_metric_levels)
  )

featured_panels <- data.frame(
  metric = c(
    "gs_max", "gs_max", "gs_max", "gs_max",
    "VPD_opt", "VPD_opt", "VPD_opt",
    "VPD50",    "VPD50",    "VPD50",
    "Sclosure", "Sclosure", "Sclosure"
  ),
  trait = c(
    "SLA", "leaf_N_area", "thick", "tough",
    "SD",  "SLA", "thick",
    "SD",  "SLA", "thick",
    "SD",  "SLA", "thick"
  )
) %>%
  mutate(
    trait  = factor(trait,  levels = correct_trait_levels),
    metric = factor(metric, levels = correct_metric_levels)
  )


overall_effect_plot <- 
  ggplot(all_draws, aes(x = slope, y = tree_group, color = tree_group, shape = tree_group)) +
  
    # Background highlight for panels with at least one credible effect
  geom_rect(
    data = highlight_panels,
    aes(fill = highlight),
    xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
    inherit.aes = FALSE,
    fill = "grey92", color = NA
  ) +
  
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60") +

  stat_pointinterval(
    .width = c(0.66, 0.95),
    point_size = 2.5,
    interval_size_range = c(0.6, 1.6), # thin (95% credible interval) vs thick (66% credible interval) line
    alpha=0.5
    ) +
  
  facet_grid(
    rows  = vars(trait),
    cols  = vars(metric),
    scales = "free_x",
    switch = "y",                          # moves row strips to the LEFT
    labeller = labeller(
      trait  = trait_labels,              # plain text names
      metric = metric_labeller            # parsed plotmath for subscripts
    )
  ) +
  
  scale_color_manual(
    values = c("broadleaved" = "#41624D", "conifer" = "#7798B9"),
    labels = c("broadleaved" = "Broadleaved", "conifer" = "Conifer")
  ) +
  scale_shape_manual(
    values = c("broadleaved" = 16, "conifer" = 17),
    labels = c("broadleaved" = "Broadleaved", "conifer" = "Conifer")
  ) +
  
  labs(
    x     = "Slope estimate",
    y     = NULL,
    color = NULL,
    shape = NULL
  ) +
  
  theme_bw(base_size = 11) +
  theme(
    strip.background  = element_blank(),
    strip.placement  = "outside",                        
    strip.text.x      = element_text(face = "bold", size = 13),
    strip.text.y.left = element_text(angle = 0, face = "bold", hjust = 1, size = 11),
    panel.grid.minor = element_blank(),
    panel.grid.major   = element_blank(),
    legend.position  = "bottom",
    axis.text.y      = element_blank(),
    axis.ticks.y     = element_blank()
  )


ggsave(here("output", "Fig_6_effect_plot.png"),
       overall_effect_plot, height = 6, width = 8, dpi=300)

ggsave(here("output", "Figure_6.pdf"),
       overall_effect_plot, height = 6, width = 8, dpi=1000)

```


## 3.2. Leaf Traits - single figures
### 3.2.1. function for figures
```{r model figures - function}

metric_labels <- list(
  "gs_max" = expression(
    paste(
      "Maximum ", italic(g)[s], " (", italic(g)[max],
      "; mol ", m^-2, " ", s^-1, ")"
    )
  ),
  "VPD_opt" = expression(
    paste("VPD optimum (VPD", scriptscriptstyle(italic(opt)), ", kPa)")
  ),
  "VPD50" = expression(
    paste("VPD tolerance (VPD", scriptscriptstyle(50), ", kPa)")
  ),
  "Sclosure" = expression(
    paste("Closure rate (", italic(S)[closure], ")")
  )
)


trait_labels <- list(
  "SD"          = expression(paste("Stomatal Density (no. ", mm^-2, ")")),
  "GCL"         = expression(paste("GCL(", mu, "m)")),
  "SLA"         = expression(paste("SLA (", m^2, " ", kg^-1, ")")),
  "LDMC"        = expression(paste("LDMC (mg ", g^-1, ")")),
  "leaf_N_area" = expression(paste("Leaf N (", g, " ", m^-2, ")")),
  "thick"       = "Leaf Thickness (mm)",
  "tough"       = expression(paste("Leaf Toughness (N ", mm^-1, ")"))
)


generate_trait_panel <- function(metric, trait, raw_data, show_y_label = TRUE) {
  
  # 1. Identify model and load it from your saved output folder
  model_name <- paste0("model_", metric, "_", trait, "_scaled")
  file_path <- here::here("output", "models", paste0("model_", metric, "_", trait, "_scaled.rds"))
  
  if (!file.exists(file_path)) {
    warning(paste("Model file not found:", file_path))
    return(ggplot() + labs(title = paste("Missing:", model_name)) + theme_void())
  }
  
  model <- readRDS(file_path)
  
  # Define variable column mappings from raw data
  metric_mean_col <- paste0(metric, "_mean")
  metric_sd_col   <- paste0(metric, "_sd")
  trait_raw_col   <- trait 
  trait_scale_col <- paste0(trait, "_scaled")
  
  # 2. Extract population levels trends & intercepts (on scaled space)
  slopes_df     <- as.data.frame(emtrends(model, ~ tree_group, var = trait_scale_col))
  colnames(slopes_df)[colnames(slopes_df) == paste0(trait_scale_col, ".trend")] <- "trend"
  
  at_list <- list(); at_list[[trait_scale_col]] <- 0
  intercepts_df <- as.data.frame(emmeans(model, ~ tree_group, at = at_list))
  
  angio_int   <- intercepts_df %>% filter(tree_group == "broadleaved") %>% pull(emmean)
  angio_slp   <- slopes_df %>% filter(tree_group == "broadleaved") %>% pull(trend)
  conifer_int <- intercepts_df %>% filter(tree_group == "conifer") %>% pull(emmean)
  conifer_slp <- slopes_df %>% filter(tree_group == "conifer") %>% pull(trend)

  
    # Summarize posterior slope distributions using 90% and 95% credible intervals.
    # These intervals determine whether relationships are classified as significant
    # or marginal.  
  slope_draws <- model %>% emtrends(~ tree_group, var = trait_scale_col) %>%
  gather_emmeans_draws()
  
  colnames(slope_draws)[colnames(slope_draws) == ".value"] <- "slope"
  
  # --- FIXED: Force .width to text to prevent R from dropping trailing zeros ---
  ci_summary <- slope_draws %>% 
    group_by(tree_group) %>% 
    tidybayes::median_qi(slope, .width = c(0.66, 0.95)) %>%
    dplyr::select(tree_group, .width, .lower, .upper) %>%
    dplyr::mutate(.width = ifelse(.width == 0.95, "95", "66")) %>% # Explicitly stringify
    tidyr::pivot_wider(
      names_from = .width, 
      values_from = c(.lower, .upper),
      names_sep = "_"
    )
  
  
  determine_status <- function(lower_95, upper_95, lower_66, upper_66) {
    if ((lower_95 > 0 && upper_95 > 0) || (lower_95 < 0 && upper_95 < 0)) {
      return("significant")
    } else if ((lower_66 > 0 && upper_66 > 0) || (lower_66 < 0 && upper_66 < 0)) {
      return("marginal")
    } else {
      return("none")
    }
  }
  
  angio_ci <- ci_summary %>% filter(tree_group == "broadleaved")
  conifer_ci <- ci_summary %>% filter(tree_group == "conifer")
  
  # Safely pull the explicitly named text columns
  angio_status <- determine_status(
    pull(angio_ci, .lower_95), pull(angio_ci, .upper_95), 
    pull(angio_ci, .lower_66), pull(angio_ci, .upper_66)
  )
  conifer_status <- determine_status(
    pull(conifer_ci, .lower_95), pull(conifer_ci, .upper_95), 
    pull(conifer_ci, .lower_66), pull(conifer_ci, .upper_66)
  )
  # -------------------------------------------------------------------------
  # -------------------------------------------------------------------------
  # -------------------------------------------------------------------------
  
  intercept_draws <- model %>% emmeans(~ tree_group, at = at_list) %>% gather_emmeans_draws()
  colnames(intercept_draws)[colnames(intercept_draws) == ".value"] <- "intercept"
  
  all_lines <- inner_join(
    intercept_draws, slope_draws,
    by = c("tree_group", ".chain", ".iteration", ".draw")
  )
  
  set.seed(123)
  random_draws <- sample(unique(all_lines$.draw), 200)
  cloud_data   <- all_lines %>% filter(.draw %in% random_draws)
  
  mean_trait <- mean(raw_data[[trait_raw_col]], na.rm = TRUE)
  sd_trait   <- sd(raw_data[[trait_raw_col]], na.rm = TRUE)
  
  cloud_data <- cloud_data %>% 
    mutate(
      slope_unscaled = slope / sd_trait,
      intercept_unscaled = intercept - (slope * mean_trait / sd_trait)
    )
  
  angio_color <- "#41624D"; angio_cloud_color = "#A9C7A9"
    conifer_color <- "#7798B9"; conifer_cloud_color <- "lightblue3"



  # --- CHANGED: Map status to linetype and filter out "none" ---
  mean_lines <- data.frame(
    tree_group = c("broadleaved", "conifer"),
    intercept_mean = c(angio_int - (angio_slp * mean_trait / sd_trait), 
                       conifer_int - (conifer_slp * mean_trait / sd_trait)),
    slope_mean = c(angio_slp / sd_trait, 
                   conifer_slp / sd_trait),
    line_color = c(angio_color, conifer_color),
    status = c(angio_status, conifer_status)
  )
  
  # Keep both significant and marginal lines
  mean_lines <- mean_lines %>% filter(status %in% c("significant", "marginal"))
  # --------------------------------------------------------------
  
  angio_cloud   <- cloud_data %>% filter(tree_group == "broadleaved")
  conifer_cloud <- cloud_data %>% filter(tree_group == "conifer")
  
  angio_raw   <- raw_data %>% filter(tree_group == "broadleaved")
  conifer_raw <- raw_data %>% filter(tree_group == "conifer")
  
  x_label <- if (trait %in% names(trait_labels)) trait_labels[[trait]] else trait
  
  y_label <- NULL
  if (show_y_label) {
    y_label <- if (metric %in% names(metric_labels)) metric_labels[[metric]] else metric
  }
  
  # Build the plot
  p <- ggplot() +
    geom_abline(data = conifer_cloud, 
                aes(intercept = intercept_unscaled, 
                    slope = slope_unscaled), 
                color = conifer_cloud_color, 
                alpha = 0.2, 
                linewidth = 0.3) + 
    geom_errorbar(data = conifer_raw, 
                  aes(x = .data[[trait_raw_col]],
                      ymin = .data[[metric_mean_col]] - .data[[metric_sd_col]], 
                      ymax = .data[[metric_mean_col]] + .data[[metric_sd_col]]),
                  color = "gray50", 
                  width = 0, 
                  alpha = 0.5) +
    geom_point(data = conifer_raw, 
               aes(x = .data[[trait_raw_col]], 
                   y = .data[[metric_mean_col]], 
                   size = 1/.data[[metric_sd_col]]),
               color = conifer_color, 
               alpha = 0.6, 
               shape = 17) +
    
    geom_abline(data = angio_cloud, 
                aes(intercept = intercept_unscaled, 
                    slope = slope_unscaled), 
                color = angio_cloud_color, 
                alpha = 0.2, 
                linewidth = 0.3) + 
    
    geom_errorbar(data = angio_raw, 
                  aes(x = .data[[trait_raw_col]], 
                      ymin = .data[[metric_mean_col]] - .data[[metric_sd_col]], 
                      ymax = .data[[metric_mean_col]] + .data[[metric_sd_col]]), 
                  color = "gray50", 
                  width = 0, 
                  alpha = 0.5) +
    
    geom_point(data = angio_raw, 
               aes(x = .data[[trait_raw_col]], 
                   y = .data[[metric_mean_col]], 
                   size = 1/.data[[metric_sd_col]]), 
               color = angio_color, 
               alpha = 0.6,
               shape = 16) +
    
    # --- CHANGED: Dynamic linetype based on status mapping ---
    geom_abline(data = mean_lines, 
                aes(intercept = intercept_mean, 
                    slope = slope_mean, 
                    color = line_color, 
                    linetype = status),
                linewidth = 1.0) +
    
    scale_linetype_manual(values = c("significant" = "solid", "marginal" = "dashed")) +
    # ---------------------------------------------------------
    scale_size_continuous(range = c(1, 4)) +
    scale_color_identity() + 
    labs(x = x_label, y = y_label) +
    theme_bw() +
    theme(
      legend.position = "none",
      strip.background = element_rect(fill = "gray95", color = NA),
      strip.text = element_text(face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    facet_wrap(~ factor(tree_group, levels = c("broadleaved", "conifer")), nrow = 1, scales = "free_x")
  
  if (!show_y_label) {
    p <- p + theme(
      axis.title.y = element_blank(),
      axis.text.y  = element_blank(),
      axis.ticks.y = element_blank(),
    )
  }
  
  return(p)
}

```

### 3.2.2. Fig. 7 - Leaf traits & Max. capacity figure

```{r gs max figure}
# Generate the individual panels
p_sla   <- generate_trait_panel(metric = "gs_max", trait = "SLA",         raw_data = data_for_models)
p_leafn <- generate_trait_panel(metric = "gs_max", trait = "leaf_N_area", raw_data = data_for_models, show_y_label =F)
p_thick <- generate_trait_panel(metric = "gs_max", trait = "thick",       raw_data = data_for_models, show_y_label =F)
p_tough <- generate_trait_panel(metric = "gs_max", trait = "tough",       raw_data = data_for_models, show_y_label =F)


fig_gsmax <- (p_sla | p_leafn | p_thick | p_tough) +
  plot_annotation(
    tag_levels = "a", 
    tag_prefix = "(", 
    tag_suffix = ")"
  ) &
   theme(
     plot.tag = element_text(face = "bold", size = 12),
     strip.text = element_text(size = 9),
     axis.title = element_text(size = 12),
     panel.spacing = unit(0.5, "lines"),
     panel.grid.major = element_blank(),
     panel.grid.minor = element_blank()
   )

fig_gsmax

ggsave(here("output", "Fig_7_gs_max.png"), fig_gsmax, height=4, width=9, dpi=300)

ggsave(here("output", "Figure_7.pdf"), fig_gsmax, height=4, width=9, dpi=1000)

```

### 3.2.3. Fig. 8 - Leaf traits & Stomatal sensitivity figure
```{r model figures - SLA}

### Row 1: VPD optimum (VPD_opt)
p_vpd_sd    <- generate_trait_panel("VPD_opt", "SD",    data_for_models, show_y_label = TRUE)
p_vpd_sla   <- generate_trait_panel("VPD_opt", "SLA",   data_for_models, show_y_label = FALSE)
p_vpd_thick <- generate_trait_panel("VPD_opt", "thick", data_for_models, show_y_label = FALSE)
p_vpd_tough <- generate_trait_panel("VPD_opt", "tough", data_for_models, show_y_label = FALSE)

### Row 2: VPD tolerance (VPD50)
p_v50_sd    <- generate_trait_panel("VPD50",    "SD",    data_for_models, show_y_label = TRUE)
p_v50_sla   <- generate_trait_panel("VPD50",    "SLA",   data_for_models, show_y_label = FALSE)
p_v50_thick <- generate_trait_panel("VPD50",    "thick", data_for_models, show_y_label = FALSE)
p_v50_tough <- generate_trait_panel("VPD50",    "tough", data_for_models, show_y_label = FALSE)

### Row 3: Closure rate (Sclosure)
p_sdec_sd    <- generate_trait_panel("Sclosure", "SD",    data_for_models, show_y_label = TRUE)
p_sdec_sla   <- generate_trait_panel("Sclosure", "SLA",   data_for_models, show_y_label = FALSE)
p_sdec_thick <- generate_trait_panel("Sclosure", "thick", data_for_models, show_y_label = FALSE)
p_sdec_tough <- generate_trait_panel("Sclosure", "tough", data_for_models, show_y_label = FALSE)

# Arrange into a 3x4 layout grid using patchwork wrapping
fig_vpd_traits <- (
  p_vpd_sd  | p_vpd_sla  | p_vpd_thick  |
  p_v50_sd  | p_v50_sla  | p_v50_thick  |
  p_sdec_sd | p_sdec_sla | p_sdec_thick 
) + 
  plot_layout(ncol = 3, nrow =3) +
  plot_annotation(
    tag_levels = "a", 
    tag_prefix = "(", 
    tag_suffix = ")"
  ) &
   theme(
     plot.tag = element_text(face = "bold", size = 12),
     strip.text = element_text(size = 9),
     axis.title = element_text(size = 10),
     panel.spacing = unit(0.5, "lines")
   )

fig_vpd_traits

ggsave(
  here::here("output", "Fig_8_vpd_metrics.png"), 
  fig_vpd_traits, 
  height = 10, 
  width = 8
)


```


## 3.3. Physiological drought response


### 3.3.1. Reload the saved models

```{r reload the models}
 
load_metric_models <- function(metrics, predictors, suffix = "") {
  for (m in metrics) {
    for (p in predictors) {
      model_name <- paste0("model_", m, "_", p, suffix)
      path <- here::here("output", "models", paste0(model_name, ".rds"))

      if (!file.exists(path)) {
        stop("Model file not found: ", path, 
             "\nRun run_metric_models() first to fit and save it.")
      }

      assign(model_name, readRDS(path), envir = .GlobalEnv)
      message("Loaded: ", model_name)
    }
  }
}

# usage in the figures script:
load_metric_models(
  metrics    = c("gs_max", "VPD_opt", "VPD50", "Sclosure"),
  predictors = "delta_d13C_scaled"
)

```

### 3.3.2. Data preparation and figure function
```{r data preparation d13C figure }

# shared styles
conifer_color       <- "#7798B9"
conifer_cloud_color <- "lightblue3"
angio_color         <- "#41624D"
angio_cloud_color   <- "#A9C7A9"



# Extract everything needed for one model: pooled fits, posterior draw
#    clouds (unscaled), Bayes R2, and slope significance text.
# ------------------------------------------------------------------------------
extract_drought_fit <- function(model, mean_d13C, sd_d13C,
                                 n_draws = 200, seed = 123) {
 
  slopes_df     <- as.data.frame(emtrends(model, ~ tree_group, var = "delta_d13C_scaled"))
  intercepts_df <- as.data.frame(emmeans(model, ~ tree_group, at = list(delta_d13C_scaled = 0)))
 
  pull_by_group <- function(df, group, col) {
    df %>% filter(tree_group == group) %>% pull({{ col }})
  }
 
  angio_int   <- pull_by_group(intercepts_df, "broadleaved", emmean)
  angio_slp   <- pull_by_group(slopes_df,     "broadleaved", delta_d13C_scaled.trend)
  conifer_int <- pull_by_group(intercepts_df, "conifer",     emmean)
  conifer_slp <- pull_by_group(slopes_df,     "conifer",     delta_d13C_scaled.trend)
 
  # Posterior draws for the uncertainty clouds
  slope_draws <- model %>%
    emtrends(~ tree_group, var = "delta_d13C_scaled") %>%
    gather_emmeans_draws()
 
  intercept_draws <- model %>%
    emmeans(~ tree_group, at = list(delta_d13C_scaled = 0)) %>%
    gather_emmeans_draws()
 
  all_lines <- inner_join(
    intercept_draws %>% rename(intercept = .value),
    slope_draws     %>% rename(slope = .value),
    by = c("tree_group", ".chain", ".iteration", ".draw")
  )
 
  set.seed(seed)
  random_draws <- sample(unique(all_lines$.draw), n_draws)
  cloud_data   <- all_lines %>% filter(.draw %in% random_draws)
 
  unscale_cloud <- function(cloud) {
    cloud %>%
      mutate(
        slope_unscaled     = slope / sd_d13C,
        intercept_unscaled = intercept - (slope * mean_d13C / sd_d13C)
      )
  }
 
  angio_cloud   <- cloud_data %>% filter(tree_group == "broadleaved") %>% unscale_cloud()
  conifer_cloud <- cloud_data %>% filter(tree_group == "conifer")     %>% unscale_cloud()
 
  # Unscale the main pooled regression lines
  angio_slp_unscaled   <- angio_slp / sd_d13C
  angio_int_unscaled   <- angio_int - (angio_slp * mean_d13C / sd_d13C)
  conifer_slp_unscaled <- conifer_slp / sd_d13C
  conifer_int_unscaled <- conifer_int - (conifer_slp * mean_d13C / sd_d13C)
 
  # Bayes R2
  r2_val  <- bayes_R2(model)[1, "Estimate"]
  r2_text <- paste0("Bayes R² = ", round(r2_val, 2))
 
  # Slope summaries and classification based on 95% and 66% HPD intervals
slope_summary_95 <- as.data.frame(
  emtrends(
    model,
    ~ tree_group,
    var = "delta_d13C_scaled",
    level = 0.95
  )
)

slope_summary_66 <- as.data.frame(
  emtrends(
    model,
    ~ tree_group,
    var = "delta_d13C_scaled",
    level = 0.66
  )
)

get_slope_status <- function(group) {
  
  ci95 <- slope_summary_95 %>%
    filter(tree_group == group)
  
  ci66 <- slope_summary_66 %>%
    filter(tree_group == group)
  
  if ((ci95$lower.HPD > 0 && ci95$upper.HPD > 0) ||
      (ci95$lower.HPD < 0 && ci95$upper.HPD < 0)) {
    "significant"
    
  } else if ((ci66$lower.HPD > 0 && ci66$upper.HPD > 0) ||
             (ci66$lower.HPD < 0 && ci66$upper.HPD < 0)) {
    "marginal"
    
  } else {
    "none"
  }
}

make_sig_text <- function(group) {
  trend <- slope_summary_95 %>%
    filter(tree_group == group)
  
  sprintf(
    "Slope = %.2f [%.2f, %.2f]",
    trend$delta_d13C_scaled.trend,
    trend$lower.HPD,
    trend$upper.HPD
  )
}

angio_status   <- get_slope_status("broadleaved")
conifer_status <- get_slope_status("conifer")
 
list(
  angio_cloud            = angio_cloud,
  conifer_cloud          = conifer_cloud,
  angio_int_unscaled     = angio_int_unscaled,
  angio_slp_unscaled     = angio_slp_unscaled,
  conifer_int_unscaled   = conifer_int_unscaled,
  conifer_slp_unscaled   = conifer_slp_unscaled,
  r2_text                = r2_text,
  angio_sig              = make_sig_text("broadleaved"),
  conifer_sig            = make_sig_text("conifer"),
  angio_status            = angio_status,
  conifer_status          = conifer_status
)
}




# figure function
build_drought_panel <- function(raw_data, fit, mean_col, sd_col, y_lab,
                                 show_conifer_fit = FALSE,
                                 show_angio_fit   = FALSE,
                                 halo             = FALSE,
                                 angio_suffix     = "",
                                 conifer_suffix   = "") {
 
  conifer_raw <- raw_data %>% filter(tree_group == "conifer")
  angio_raw   <- raw_data %>% filter(tree_group == "broadleaved")
 
  p <- ggplot() +
    geom_abline(data = fit$conifer_cloud,
                aes(intercept = intercept_unscaled, slope = slope_unscaled),
                color = conifer_cloud_color, alpha = 0.3, linewidth = 0.3) +
    geom_abline(data = fit$angio_cloud,
                aes(intercept = intercept_unscaled, slope = slope_unscaled),
                color = angio_cloud_color, alpha = 0.3, linewidth = 0.3) +
    geom_errorbar(data = conifer_raw,
                  aes(x = delta_d13C,
                      ymin = .data[[mean_col]] - .data[[sd_col]],
                      ymax = .data[[mean_col]] + .data[[sd_col]]),
                  color = "gray50", width = 0, alpha = 0.6) +
    geom_errorbar(data = angio_raw,
                  aes(x = delta_d13C,
                      ymin = .data[[mean_col]] - .data[[sd_col]],
                      ymax = .data[[mean_col]] + .data[[sd_col]]),
                  color = "gray50", width = 0, alpha = 0.6)
 

# White halo only behind 95% credible (solid) fit lines
if (halo) {
  
  if (show_conifer_fit && fit$conifer_status == "significant") {
    p <- p + geom_abline(
      intercept = fit$conifer_int_unscaled,
      slope = fit$conifer_slp_unscaled,
      color = "white",
      linewidth = 2.0,
      linetype = "solid"
    )
  }
  
  if (show_angio_fit && fit$angio_status == "significant") {
    p <- p + geom_abline(
      intercept = fit$angio_int_unscaled,
      slope = fit$angio_slp_unscaled,
      color = "white",
      linewidth = 2.0,
      linetype = "solid"
    )
  }
}
 
  p <- p +
    geom_point(data = conifer_raw,
               aes(x = delta_d13C, y = .data[[mean_col]], size = 1 / .data[[sd_col]]),
               color = conifer_color, alpha = 0.8, shape = 17) +
    geom_point(data = angio_raw,
               aes(x = delta_d13C, y = .data[[mean_col]], size = 1 / .data[[sd_col]]),
               color = angio_color, alpha = 0.8)
 
    if (show_conifer_fit && fit$conifer_status != "none") {
      p <- p + geom_abline(
        intercept = fit$conifer_int_unscaled,
        slope = fit$conifer_slp_unscaled,
        color = conifer_color,
        linewidth = 1.2,
        linetype = ifelse(
          fit$conifer_status == "marginal",
          "dashed",
          "solid"
        )
      )
    }
    
    if (show_angio_fit && fit$angio_status != "none") {
      p <- p + geom_abline(
        intercept = fit$angio_int_unscaled,
        slope = fit$angio_slp_unscaled,
        color = angio_color,
        linewidth = 1.2,
        linetype = ifelse(
          fit$angio_status == "marginal",
          "dashed",
          "solid"
        )
      )
    }
 
  p +
    annotate("text", x = Inf, y = Inf, label = fit$r2_text,
             hjust = 1, vjust = 1.5, color = "black", fontface = "plain", size = 4) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("Broadl.: ", fit$angio_sig, angio_suffix),
             hjust = 1, vjust = 3.2, color = angio_color, fontface = "plain", size = 3.8) +
    annotate("text", x = Inf, y = Inf,
             label = paste0("Conifer: ", fit$conifer_sig, conifer_suffix),
             hjust = 1, vjust = 4.9, color = conifer_color, fontface = "plain", size = 3.8) +
    labs(
      x = expression(atop("Physiol. drought response",
        (Delta * delta^13 * "C"["2018-2021"]))),
      y = y_lab
    ) +
    theme_minimal() +
    theme(legend.position = "none")
}



```

### 3.3.3. Fig. 5 - Stomatal behaviour & drought response
```{r figures d13C}

mean_d13C <- mean(data_for_models$delta_d13C, na.rm = TRUE)
sd_d13C   <- sd(data_for_models$delta_d13C, na.rm = TRUE)
 
fit_gsmax      <- extract_drought_fit(model_gs_max_delta_d13C_scaled,      mean_d13C, sd_d13C)
fit_VPDoptimum <- extract_drought_fit(model_VPD_opt_delta_d13C_scaled, mean_d13C, sd_d13C)
fit_VPD50      <- extract_drought_fit(model_VPD50_delta_d13C_scaled,      mean_d13C, sd_d13C)
fit_Sclosure   <- extract_drought_fit(model_Sclosure_delta_d13C_scaled,   mean_d13C, sd_d13C)
 
 
# gs_max: no fitted lines drawn, no significance suffixes
figure_Dd13C_gsmax <- build_drought_panel(
  data_for_models, fit_gsmax, "gs_max_mean", "gs_max_sd",
  y_lab = expression(
    paste(
      "Maximum ", italic(g)[s], " (", italic(g)[max],
      "; mol ", m^-2, " ", s^-1, ")"
    )
  ),
  show_conifer_fit = TRUE,
  show_angio_fit   = TRUE,
  halo             = FALSE
)

# VPDoptimum: both fit lines shown, with white halo, both starred
figure_Dd13C_VPDopt <- build_drought_panel(
  data_for_models, fit_VPDoptimum, "VPD_opt_mean", "VPD_opt_sd",
  y_lab            = expression("VPD optimum (VPD"["opt"]*")"),
  show_conifer_fit = TRUE,
  show_angio_fit   = TRUE,
  halo             = TRUE,
  angio_suffix     = "*",
  conifer_suffix   = "*"
)
 
# VPD50: only conifer fit line shown (with halo), conifer starred
figure_Dd13C_VPD50 <- build_drought_panel(
  data_for_models, fit_VPD50, "VPD50_mean", "VPD50_sd",
  y_lab            = expression("VPD at half-closure (VPD"["50"]*")"),
  show_conifer_fit = TRUE,
  show_angio_fit   = TRUE,
  halo             = TRUE,
  angio_suffix     = " ",
  conifer_suffix   = "*"
)
 
# Sclosure: no fitted lines, no suffixes
figure_Dd13C_Sclosure <- build_drought_panel(
  data_for_models, fit_Sclosure, "Sclosure_mean", "Sclosure_sd",
  y_lab = expression("Closure rate (S"["closure"]*")"),
  show_conifer_fit = TRUE,
  show_angio_fit   = TRUE,
  halo             = TRUE,
)
 

# Combined drought figure

fig_Dd13C <- (
  figure_Dd13C_gsmax +
  figure_Dd13C_VPDopt +
  figure_Dd13C_VPD50 +
  figure_Dd13C_Sclosure
) +
  plot_layout(nrow = 1) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  ) &
  theme(
    plot.tag  = element_text(size = 18, face = "bold"),
    axis.title = element_text(size = 18),
    axis.text  = element_text(size = 12),
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.spacing = unit(0.5, "lines"),
  )
fig_Dd13C
 
ggsave(
  here("output", "Fig_5_Model_Dd13C.png"),
  fig_Dd13C, height = 7, width = 15, dpi=300
)

ggsave(
  here("output", "Figure_5.pdf"),
  fig_Dd13C, height = 7, width = 15, dpi=1000
)




```




# 4. Supplementary Figures

## 4.1. Fig. S11 - PCA Leaf Traits 

```{r pca leaf traits}

tree_group_info <- data_mean %>% dplyr::select(species, tree_group) %>% distinct() %>% ungroup()

# only angio

data_pca <- data_mean %>%
  filter(tree_group=="broadleaved") %>% 
  dplyr::select(SD, GCL, SLA, LDMC, leaf_N_area, thick, tough) %>% 
  as.data.frame()

species_names <- data_mean %>% 
  filter(tree_group=="broadleaved") %>% 
  pull(species)

rownames(data_pca) <- species_names

pca_res <- prcomp(data_pca, 
                  center = TRUE, 
                  scale. = TRUE)


# 1. Extract Individual Scores (the points)
ind_coords <- as.data.frame(pca_res$x)
ind_coords$species <- rownames(ind_coords)
ind_coords <- ind_coords %>% left_join(tree_group_info)

# 2. Extract Variable Loadings (the arrows)
# We multiply by a scaling factor so arrows are visible on the same scale as points
var_coords <- as.data.frame(pca_res$rotation)
var_coords$var_names <- rownames(var_coords)

# flip both axes for angio so trait directions match gymno's orientation
ind_coords$PC1 <- -ind_coords$PC1
ind_coords$PC2 <- -ind_coords$PC2
var_coords$PC1 <- -var_coords$PC1
var_coords$PC2 <- -var_coords$PC2

# extract the scores
pca_scores_angio <- ind_coords %>%
  dplyr::select(species, PC1, PC2)

mult <- min(
  (max(ind_coords$PC1) - min(ind_coords$PC1)) /
    (max(var_coords$PC1) - min(var_coords$PC1)),
  (max(ind_coords$PC2) - min(ind_coords$PC2)) /
    (max(var_coords$PC2) - min(var_coords$PC2))
) * 0.7 # Adjust this multiplier to change arrow length

# 1. Calculate percentage of variance
# sdev is the square root of the eigenvalues
pca_var <- (pca_res$sdev^2) / sum(pca_res$sdev^2) * 100
pc1_lab <- paste0("PC1 (", round(pca_var[1], 1), "%)")
pc2_lab <- paste0("PC2 (", round(pca_var[2], 1), "%)")

pca_traits_angio <- ggplot(ind_coords, aes(x = PC1, y = PC2)) +
      # Styling to match fviz_pca
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  # Variable arrows
  geom_segment(data = var_coords, 
               aes(x = 0, y = 0, xend = PC1 * mult, yend = PC2 * mult),
               arrow = arrow(length = unit(0.2, "cm")), 
               color = "black", linewidth = 0.7) +
  
  # Variable labels
  geom_text_repel(data = var_coords, 
                  aes(x = PC1 * mult, y = PC2 * mult, label = var_names),
                  color = "black", fontface = "bold", size = 4) +
  
  # Individual points & labels
  geom_point(aes(color = tree_group, shape=tree_group), size = 3) +
  geom_text_repel(aes(label = species, colour=tree_group), size = 3) +
    
  scale_colour_manual(
    values = c(
      "broadleaved" = "#41624D",
      "conifer" = "#7798B9"
    )
  ) +
    scale_shape_manual(
    values = c(
      "broadleaved" = 16,
      "conifer" = 17
    )
  ) +
theme_minimal() +
labs(x = pc1_lab, y = pc2_lab) +
# coord_cartesian(xlim = x_limits, ylim = y_limits) +
coord_fixed() +
theme(legend.position = "none")
pca_traits_angio



# only gymno

data_pca <- data_mean %>%
  filter(tree_group=="conifer") %>% 
  dplyr::select(SD, GCL, SLA, LDMC, leaf_N_area, thick, tough) %>% 
  as.data.frame()

species_names <- data_mean %>% 
  filter(tree_group=="conifer") %>%
  pull(species)

rownames(data_pca) <- species_names

pca_res <- prcomp(data_pca, 
                  center = TRUE, 
                  scale. = TRUE)


# 1. Extract Individual Scores (the points)
ind_coords <- as.data.frame(pca_res$x)
ind_coords$species <- rownames(ind_coords)
ind_coords <- ind_coords %>% left_join(tree_group_info)

# extract the scores
pca_scores_gymno <- ind_coords %>%
  dplyr::select(species, PC1, PC2)


# 2. Extract Variable Loadings (the arrows)
# We multiply by a scaling factor (e.g., 5) so arrows are visible on the same scale as points
var_coords <- as.data.frame(pca_res$rotation)
var_coords$var_names <- rownames(var_coords)
mult <- min(
  (max(ind_coords$PC1) - min(ind_coords$PC1)) /
    (max(var_coords$PC1) - min(var_coords$PC1)),
  (max(ind_coords$PC2) - min(ind_coords$PC2)) /
    (max(var_coords$PC2) - min(var_coords$PC2))
) * 0.7 # Adjust this multiplier to change arrow length





# 1. Calculate percentage of variance
# sdev is the square root of the eigenvalues
pca_var <- (pca_res$sdev^2) / sum(pca_res$sdev^2) * 100
pc1_lab <- paste0("PC1 (", round(pca_var[1], 1), "%)")
pc2_lab <- paste0("PC2 (", round(pca_var[2], 1), "%)")



pca_traits_gymno <- ggplot(ind_coords, aes(x = PC1, y = PC2)) +
  
  # Styling to match fviz_pca
    geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
    # Variable arrows
  geom_segment(data = var_coords, 
               aes(x = 0, y = 0, xend = PC1 * mult, yend = PC2 * mult),
               arrow = arrow(length = unit(0.2, "cm")), 
               color = "black", linewidth = 0.7) +
  
  # Variable labels
  geom_text_repel(data = var_coords, 
                  aes(x = PC1 * mult, y = PC2 * mult, label = var_names),
                  color = "black", fontface = "bold", size = 4) +
  # Individual points & labels
  geom_point(aes(color = tree_group, shape=tree_group), size = 3) +
  geom_text_repel(aes(label = species, colour=tree_group), size = 3) +
    
  scale_colour_manual(
    values = c(
      "broadleaved" = "#41624D",
      "conifer" = "#7798B9"
    )
  ) +
    scale_shape_manual(
    values = c(
      "broadleaved" = 16,
      "conifer" = 17
    )
  ) +
theme_minimal() +
labs(x = pc1_lab, y = pc2_lab) +
# coord_cartesian(xlim = x_limits, ylim = y_limits) +
coord_fixed() +
theme(legend.position = "none")

pca_traits_gymno


# joined figure
PCA_all_leaf_traits_separated <- 
  pca_traits_angio +
  pca_traits_gymno +
  plot_layout(nrow(1))

ggsave(here("output", "Figure_S11_PCA_leaf_traits.png"), PCA_all_leaf_traits_separated, height=7, width=11)


# extract the PCA scors
pca_scores_separate <- bind_rows(
  pca_scores_angio,
  pca_scores_gymno
)



# PCA overall

data_pca <- data_mean %>%
  dplyr::select(SD, GCL, SLA, LDMC, leaf_N_area, thick, tough) %>% 
  as.data.frame()

species_names <- data_mean %>% 
  pull(species)

rownames(data_pca) <- species_names

pca_res <- prcomp(data_pca, 
                  center = TRUE, 
                  scale. = TRUE)


# 1. Extract Individual Scores (the points)
ind_coords <- as.data.frame(pca_res$x)
ind_coords$species <- rownames(ind_coords)
ind_coords <- ind_coords %>% left_join(tree_group_info)


# 2. Extract Variable Loadings (the arrows)
# We multiply by a scaling factor (e.g., 5) so arrows are visible on the same scale as points
var_coords <- as.data.frame(pca_res$rotation)
var_coords$var_names <- rownames(var_coords)
mult <- min(
  (max(ind_coords$PC1) - min(ind_coords$PC1)) /
    (max(var_coords$PC1) - min(var_coords$PC1)),
  (max(ind_coords$PC2) - min(ind_coords$PC2)) /
    (max(var_coords$PC2) - min(var_coords$PC2))
) * 0.7 # Adjust this multiplier to change arrow length


# 1. Calculate percentage of variance
# sdev is the square root of the eigenvalues
pca_var <- (pca_res$sdev^2) / sum(pca_res$sdev^2) * 100
pc1_lab <- paste0("PC1 (", round(pca_var[1], 1), "%)")
pc2_lab <- paste0("PC2 (", round(pca_var[2], 1), "%)")



pca_traits_both <- ggplot(ind_coords, aes(x = PC1, y = PC2)) +
  
  # Styling to match fviz_pca
    geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
    # Variable arrows
  geom_segment(data = var_coords, 
               aes(x = 0, y = 0, xend = PC1 * mult, yend = PC2 * mult),
               arrow = arrow(length = unit(0.2, "cm")), 
               color = "black", linewidth = 0.7) +
  
  # Variable labels
  geom_text_repel(data = var_coords, 
                  aes(x = PC1 * mult, y = PC2 * mult, label = var_names),
                  color = "black", fontface = "bold", size = 4) +
  # Individual points & labels
  geom_point(aes(color = tree_group, shape=tree_group), size = 3) +
  geom_text_repel(aes(label = species, colour=tree_group), size = 3) +
    
  scale_colour_manual(
    values = c(
      "broadleaved" = "#41624D",
      "conifer" = "#7798B9"
    )
  ) +
    scale_shape_manual(
    values = c(
      "broadleaved" = 16,
      "conifer" = 17
    )
  ) +
theme_minimal() +
labs(x = pc1_lab, y = pc2_lab) +
# coord_cartesian(xlim = x_limits, ylim = y_limits) +
coord_fixed() +
theme(legend.position = "none")

pca_traits_both


# extract the scores
pca_scores_one_pca <- ind_coords %>%
  dplyr::select(species, PC1, PC2)

```



## 4.2. Fig. S12 - Stomata Density x Size

```{r density size}

# full dataset model
mod_all <- lm(GCL ~ SD, data = data_mean)
r2_all  <- summary(mod_all)$r.squared
p_all   <- summary(mod_all)$coefficients[2, 4]

# conifer-only model
mod_con <- lm(GCL ~ SD, data = data_mean %>% filter(tree_group == "conifer"))
r2_con  <- summary(mod_con)$r.squared
p_con   <- summary(mod_con)$coefficients[2, 4]

# just for having p < 0.001

format_p <- function(p) {
  if (p < 0.001) {
    "p < 0.001"
  } else if (p < 0.01) {
    "p < 0.01"
  } else if (p < 0.05) {
    "p < 0.05"
  } else {
    "p ≥ 0.05"
  }
}

p_all_txt <- format_p(p_all)
p_con_txt <- format_p(p_con)


# create labels
lab_all <- paste0("R² = ", round(r2_all, 2),
                  "\n", p_all_txt)

lab_con <- paste0("R² = ", round(r2_con, 2),
                  "\n", p_con_txt)

# correlation between SD and GCL

fig_SD_GCL_all <- 
  data_mean %>%
  ggplot(aes(x = SD, y = GCL)) +

  # vertical SE (GCL)
  geom_errorbar(
    aes(ymin = GCL - GCL_se, 
        ymax = GCL + GCL_se,
        colour = tree_group),
    width = 0,
  alpha = 0.6,
  linewidth = 0.4
  ) +

  # horizontal SE (SD) 
  geom_errorbar(
    aes(xmin = SD - SD_se, 
        xmax = SD + SD_se,
        colour=tree_group),
    height = 0,
    orientation = "y",
  alpha = 0.6,
  linewidth = 0.4
  ) +

  geom_point(aes(shape=tree_group, colour=tree_group), alpha=0.8, size = 4) +
  
  geom_smooth(method = "lm", se = FALSE, colour="darkgrey") +

  scale_shape_manual(values = c("broadleaved" = 16, "conifer" = 17)) +
  
  scale_colour_manual(
    values = c("broadleaved" = "#41624D", "conifer" = "#7798B9")
) +
annotate("text",
         x = -Inf, y = -Inf,
         label = lab_all,
         hjust = -0.1, 
         vjust = -0.5,
         size = 4) +
  theme_classic() +
  labs(
    x = "Stomatal density [N/mm²]",
    y = "Guard cell length [µm]",
    shape = "Tree Group",
    colour = "Tree Group"
  )
fig_SD_GCL_all


# inset for gymnosperms only
fig_SD_GCL_conifers <- data_mean %>%
  filter(tree_group == "conifer") %>%
  ggplot(aes(x = SD, 
             y = GCL)) +
  geom_point(shape = 17, 
             size = 3, 
             alpha = 0.7,
             colour="#7798B9") +
  geom_smooth(method = "lm", se = FALSE, colour = "darkgrey") +
  theme_classic() +
  labs(x = NULL, y = NULL) +
   annotate("text",
           x = Inf, y = Inf,
           label = lab_con,
           hjust = 1.1, vjust = 1.5,
           size = 3) +
  theme(axis.text = element_text(size = 6),
        axis.title = element_text(size = 7))


# combine main plot + inset
fig_SD_GCL <- fig_SD_GCL_all + 
  inset_element(fig_SD_GCL_conifers, 
                left = 0.55, 
                bottom = 0.55, 
                right = 0.98, 
                top = 0.98)
fig_SD_GCL

ggsave(here("output", "Fig_S12_stom_dens_size.png"), fig_SD_GCL,  height=5, width=7, dpi=300)


```



# 5. Supplementary Tables
## 5.1. Tab. S9 - Leaf trait model summary
```{r leaf trait model results table}


metrics <- c("gs_max", "VPD_opt", "VPD50", "Sclosure")
traits  <- c("SD_scaled", "GCL_scaled", "SLA_scaled", "LDMC_scaled", 
             "thick_scaled", "tough_scaled", "leaf_N_area_scaled")

summary_list_trait_models <- list()
model_counter <- 1 # Initialize the 1 to 28 numbering counter

for (m in metrics) {
  for (t in traits) {
    
    # This matches the original file name where the model is stored
    original_model_name <- paste0("model_", m, "_", t)
    file_path <- here::here("output", "models", paste0(original_model_name, ".rds"))
    
    if (!file.exists(file_path)) next
    
    fit <- readRDS(file_path)
    fit_summary <- summary(fit)$fixed
    
    # Calculate Bayes R2
    r2_val <- bayes_R2(fit)[1, "Estimate"]
    
    ref_slope_row <- t
    interaction_row <- grep(paste0("^", t, ":tree_group"), rownames(fit_summary), value = TRUE)
    
    # Clean up names for the table presentation
    clean_metric <- gsub("_scaled", "", m)
    clean_trait  <- gsub("_scaled", "", t)
    display_model_name <- paste0("Model ", model_counter)
    
    # --- Group 1: Broadleaved (The Reference Group) ---
    if (ref_slope_row %in% rownames(fit_summary)) {
      g1_stats <- fit_summary[ref_slope_row, ]
      
      summary_list_trait_models[[length(summary_list_trait_models) + 1]] <- data.frame(
        model_name = display_model_name,
        metric     = clean_metric,
        trait      = clean_trait,
        tree_group = "broadleaved",
        R2         = r2_val,
        slope      = as.numeric(g1_stats["Estimate"]),
        CI_lower   = as.numeric(g1_stats["l-95% CI"]),
        CI_upper   = as.numeric(g1_stats["u-95% CI"]),
        Rhat       = as.numeric(g1_stats["Rhat"]),
        ESS_Bulk   = as.numeric(g1_stats["Bulk_ESS"]),
        ESS_Tail   = as.numeric(g1_stats["Tail_ESS"]),
        stringsAsFactors = FALSE
      )
    }
    
    # --- Group 2: Conifers (The Interaction Group) ---
    if (length(interaction_row) > 0 && interaction_row %in% rownames(fit_summary)) {
      g2_diff  <- fit_summary[interaction_row, ]
      
      draws <- as.data.frame(fit)
      b_ref <- draws[[paste0("b_", ref_slope_row)]]
      b_int <- draws[[paste0("b_", interaction_row)]]
      g2_slope_posterior <- b_ref + b_int
      
      summary_list_trait_models[[length(summary_list_trait_models) + 1]] <- data.frame(
        model_name = display_model_name,
        metric     = clean_metric,
        trait      = clean_trait,
        tree_group = "conifer",
        R2         = r2_val,
        slope      = mean(g2_slope_posterior),
        CI_lower   = quantile(g2_slope_posterior, probs = 0.025),
        CI_upper   = quantile(g2_slope_posterior, probs = 0.975),
        Rhat       = as.numeric(g2_diff["Rhat"]),     
        ESS_Bulk   = as.numeric(g2_diff["Bulk_ESS"]), 
        ESS_Tail   = as.numeric(g2_diff["Tail_ESS"]), 
        stringsAsFactors = FALSE
      )
    }
    
    # Advance the counter for the next model combination
    model_counter <- model_counter + 1
  }
}

# Combine and apply rounding rules
summary_table_trait_models <- 
  bind_rows(summary_list_trait_models)  %>% 
  mutate(
    R2       = round(R2, 2),
    slope    = round(slope, 3),
    CI_lower = round(CI_lower, 2),
    CI_upper = round(CI_upper, 2),
    Rhat     = round(Rhat, 2),
    ESS_Bulk = round(ESS_Bulk, 0),
    ESS_Tail = round(ESS_Tail, 0)
  )

# Save files
write_xlsx(summary_table_trait_models, here("output", "Table_S9_model_summary_table_traits.xlsx"))

```


## 5.2. Tab. S10 - Leaf ∆δ13C model summary
```{r Dd13C model results table}

model_list_Dd13C <- list(
  "gsmax"   = model_gs_max_delta_d13C_scaled,
  "VPDopt"   = model_VPD_opt_delta_d13C_scaled,
  "VPD50"    = model_VPD50_delta_d13C_scaled,
  "Sclosure" = model_Sclosure_delta_d13C_scaled
)

group_levels <- levels(data_for_models$tree_group)
broadleaved <- group_levels[1]
conifer <- group_levels[2]

extract_model_metrics_named <- function(model_name, model_obj, g1, g2) {
  
  sum_mod <- summary(model_obj)
  fixed_eff <- as.data.frame(sum_mod$fixed)
  
  rhat_val <- fixed_eff["delta_d13C_scaled", "Rhat"]
  bulk_ess <- fixed_eff["delta_d13C_scaled", "Bulk_ESS"]
  tail_ess <- fixed_eff["delta_d13C_scaled", "Tail_ESS"]
  r2_val   <- bayes_R2(model_obj)[1, "Estimate"]
  
  # Calculate Slopes
  hyp_lvl1 <- hypothesis(model_obj, "delta_d13C_scaled = 0")
  
  int_term <- rownames(fixed_eff)[grep("delta_d13C_scaled:", rownames(fixed_eff))]
  hyp_formula_lvl2 <- paste("delta_d13C_scaled +", int_term, "= 0")
  hyp_lvl2 <- hypothesis(model_obj, hyp_formula_lvl2)
  
  predictor_metric <- str_remove(model_name, "model_Dd13C_")
  
  # Build the tibble row
  res <- tibble(
    Model_Name   = model_name,
    Response     = "Dd13C",
    Predictor    = predictor_metric,
    Rhat         = round(rhat_val, 3),
    ESS_Bulk     = round(bulk_ess, 0),
    ESS_Tail     = round(tail_ess, 0),
    
    # Dynamic slope columns using your group names
    !!paste0("Slope_", g1) := round(hyp_lvl1$hypothesis$Estimate, 3),
    !!paste0("CI_", g1)    := paste0("[", round(hyp_lvl1$hypothesis$CI.Lower, 3), ", ", round(hyp_lvl1$hypothesis$CI.Upper, 3), "]"),
    
    !!paste0("Slope_", g2) := round(hyp_lvl2$hypothesis$Estimate, 3),
    !!paste0("CI_", g2)    := paste0("[", round(hyp_lvl2$hypothesis$CI.Lower, 3), ", ", round(hyp_lvl2$hypothesis$CI.Upper, 3), "]"),
    
    Bayes_R2     = round(r2_val, 3)
  )
  
  return(res)
}

# 3. Run the loop across your model list
summary_table_Dd13C <- imap_dfr(model_list_Dd13C, ~extract_model_metrics_named(.y, .x, broadleaved, conifer))

write_xlsx(summary_table_Dd13C, here("output", "Table_S10_model_summary_table_Dd13C.xlsx"))

```

