---
title: "Stomatal Behaviour 2026 - 04 Environmental Conditions"
author: "Lena Sachsenmaier"
output:
  html_document:
    df_print: paged
contact: lena.sachsenmaier@idiv.de
update: "2026-07-27"
updated: "`r format(Sys.time(), '%d %B, %Y')`"
---

**updated by**: Lena Sachsenmaier
**Contact**: lena.sachsenmaier@idiv.de  
**Last updated**: `r format(Sys.time(), '%d %B, %Y')`


Loading the packages needed:

```{r setup, warning=F}

library(tidyverse)    # data manipulation and visualisation
library(here)         # project-relative file paths
library(RColorBrewer) # colour palettes for figures
library(ggpubr)       # arranging multiple ggplots
library(patchwork)    # combining ggplots
library(lme4)         # linear mixed-effects models
library(emmeans)      # estimated marginal means and trends
library(lubridate)    # working with dates and times
library(SPEI)         # SPEI drought index and PET calculations
library(soilDB)       # Rosetta pedotransfer model
library(terra)        # for work with gridded data

Sys.setlocale("LC_TIME", "C")  # use English month and weekday names

```


# 1. SPEI - Drought Index

SPEI is the standardized precipitation evapotranspiration index, a commonly used drought index.
(for more information, see Vicente-Serrano et al., 2010 https://doi.org/10.1175/2009JCLI2909.1)

We use it here to identify which years have been drought years at our site.

For the calculation of the SPEI, the R package spei() is used. See Vignette here: https://cran.r-project.org/web/packages/SPEI/SPEI.pdf 

## 1.1. Preparation of data frame

Automatized download the data from the German Meteorological Service - DWD (Deutscher Wetterdienst):
for the climate station Leipzig/Halle Airport (Station ID: 02932) and cleaning it.


```{r load the DWD data}

# 1. Set up paths and station info
station_id <- "02932"  # Leipzig/Halle
base_url <- "https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/daily/kl/historical/"
data_dir <- "data"
zip_path <- file.path(data_dir, paste0("climate_", station_id, ".zip"))

if (!dir.exists(data_dir)) dir.create(data_dir)

# 2. Get the directory listing and find the correct zip file name
# (the file name includes the date range, which changes as DWD updates it,
# so we can't hardcode it -- we scrape the index page instead)
index_page <- read_lines(base_url)
zip_filename <- str_subset(index_page, paste0("tageswerte_KL_", station_id, "_.*_hist\\.zip")) %>%
  str_extract(paste0("tageswerte_KL_", station_id, "_[0-9]+_[0-9]+_hist\\.zip")) %>%
  unique() %>%
  na.omit() %>% 
  .[1]

file_url <- paste0(base_url, zip_filename)

# 3. Download the zip file (only if not already downloaded)
if (!file.exists(zip_path)) {
  download.file(file_url, destfile = zip_path, mode = "wb")
}

# 4. Unzip into the data folder
unzip(zip_path, exdir = data_dir)

# 5. Find and read the actual product file (name starts with "produkt_klima_tag_")
climate_file <- list.files(data_dir, pattern = "^produkt_klima_tag_.*\\.txt$", full.names = TRUE)[1]
data_2932 <- read_delim(climate_file)

glimpse(data_2932)


# clean and prepare the dataframe
climate <- data_2932 %>%
  rename(
    date = MESS_DATUM,
    # wind
    wind_gust_max = `  FX`,   # max wind speed (m/s)
    wind_speed = `  FM`,      # mean wind speed (m/s)
    
    # precipitation
    precipitation = ` RSK`,   # daily precipitation (mm)
    precip_type = RSKF,       # precipitation type (code)
    
    # radiation / sunshine
    sunshine_hours = ` SDK`,  # sunshine duration (h)
    
    # cloudiness
    cloud_cover = `  NM`,     # cloud cover (1/8)
    
    # pressure & humidity
    vapour_pressure = ` VPM`, # vapor pressure (hPa)
    air_pressure = `  PM`,    # air pressure (hPa)
    rel_humidity = ` UPM`,    # relative humidity (%)
    
    # temperature
    t_mean = ` TMK`,          # daily mean temperature (°C)
    t_min = ` TNK`,           # daily min 2m temp (°C)
    t_max = ` TXK`,           # daily max 2m temp (°C)
    t_ground_min = ` TGK`,    # min temp at 5cm (°C)
    
    # snow
    snow_depth = SHK_TAG      # snow height (cm)
  ) %>%
  # convert numeric columns from character to numeric
  mutate(across(c(wind_gust_max, wind_speed, precipitation, sunshine_hours,
                  cloud_cover, vapour_pressure, air_pressure, rel_humidity,
                  t_mean, t_min, t_max, t_ground_min, snow_depth), 
                ~ as.numeric(trimws(.)))) %>% 
  # convert cloud cover in okta (1/8) to percent (%): cc % = (cc/8)*100
  mutate(cloud_cover = (cloud_cover/8)*100) %>% 
  # convert air pressure and vapor pressure from hPa to kPa: p(kPa) = p(hPa)*0.1
  mutate(air_pressure = air_pressure*0.1,
         vapour_pressure = vapour_pressure*0.1) %>% 
  mutate(date= ymd(date),
         year= year(date), # create new variable: extract year
         month=month(date)) %>% # create new variable: extract month
  filter(year>=1973) %>%  # exclude everything before 1973, because data not complete
  filter(year < 2026) %>%  # only until year 2025, because 2026 not yet complete
  arrange(date) %>% 
  relocate(year, .after=date) %>% 
  relocate(month, .after=year)



# First convert all "-999" (or strings of -999) to NA
climate <- climate %>%
  mutate(across(where(is.numeric), ~na_if(., -999)))  # replace -999 with NA

# check for NA values in different all my variables  
na_count <-sapply(climate, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)
na_count

```


## 1.2. Long-term means

```{r long-term means}

# What is the mean temperature between 1973 and 2025
climate %>% dplyr::summarize(t_mean = mean(t_mean)) # 9.7 ° C

# What is the mean annual precipitation between 1973 and 2025
climate %>% 
  group_by(year) %>% 
  summarize(annual_precip = sum(precipitation)) %>% 
  summarize(mean_annual_precip = mean(annual_precip))
            # 519.52 mm per year

```


## 1.3. Calculation of monthly means
```{r monthly means calculation}

# mothly means
monthly_climate <- climate %>%
  group_by(year, month) %>%
  summarise(
    # sum precipitation
    precipitation = sum(precipitation, na.rm = TRUE),
    
    # mean of all other numeric variables except precipitation
    across(c(t_mean, t_min, t_max, wind_gust_max, wind_speed, sunshine_hours, snow_depth, 
             cloud_cover, vapour_pressure, air_pressure, rel_humidity), 
           ~mean(.x, na.rm = TRUE))
  ) %>%
  ungroup()

```

## 1.4. Site latitude and longitude
```{r lat and long}
# latitude and longitude of the climate station Leipzig/Halle Airport in Germany
lat_station <- 51.43479
elevation_station <- 131.0 

```

## 1.5. Calculation of PET

Potential evapotranspirtation (PET) is the amount of evaporation and transpiration that would occur if a sufficient water source were available.  
There are three different ways how to calculate PET:  

* thornthwaite(Tave, lat, na.rm = FALSE, verbose=TRUE)
* hargreaves(Tmin, Tmax, Ra = NULL, lat = NULL, Pre = NULL, na.rm = FALSE,
verbose=TRUE)
* penman(Tmin, Tmax, U2, Ra = NULL, lat = NULL, Rs = NULL, tsun = NULL,
CC = NULL, ed = NULL, Tdew = NULL, RH = NULL, P = NULL, P0 = NULL,
CO2 = NULL, z = NULL, crop='short', na.rm = FALSE, method='ICID',
verbose=TRUE)

* thornthwaite computes the monthly potential evapotranspiration (PE) according to the Thornth-
waite (1948) equation. It is the simplest of the three methods, and can be used when only mean
temperature data are available. But we have more available.
* hargreaves computes the monthly reference evapotranspiration (ETo) of a grass crop based on
the original Hargreaves equation (1994). However, if precipitation data Pre is provided a modified
form due to Droogers and Allen (2002) will be used; this equation corrects ETo using the amount
of rain of each month as a proxy for irradiation The Hargreaves method requires data on the mean
external radiation, Ra. If Ra is not available it can be estimated from the latitude lat and the month
of the year.
* penman calculates the monthly reference evapotranspiration (ETo) of a hypothetical reference crop
according to the Penman-Monteith equation (Monteith, 1965). This is widely considered the most
accurate method for estimating ETo, and is the method recommended by the UN Food and Agricul-
ture Organization (FAO). There are several methods which simplify the original Penman-Monteith
equation and are used in practice. Here we follow by default the procedure described in Allen et al.
(1994), aka the ‘ICID‘ method. However, other versions are also implemented and can be used by
setting the method parameter to the appropriate value. The FAO-56 method (Allen et al., 1998) can
be used by setting method to ‘FAO‘, while the variation of the American Society of Civil Engineers
(Walter et al., 2002 is used when setting it to ‘ASCE‘. By default the original parameterization
corresponding to a short reference crop of 0.12 m height is used, although the parameterization for
a tall reference crop of 0.5 m height (Walter et al. (2002) can also be used, by setting the crop pa-
rameter to ’tall’. The method requires data on the incoming solar radiation, Rs; since this is seldom
available, the code will estimate it from data on the bright sunshine duration tsun, or alternatively
from data on the percent cloud cover CC. Similarly, if data on the saturation water pressure ed are not
available, it is possible to estimate it from the dew point temperature Tdew, from the relative humid-
ity RH or even from the minimum temperature Tmin (sorted from least to most uncertain method).
Similarly, the atmospheric surface pressure P required for computing the psychrometric constant
can be calculated from the atmospheric pressure at sea level P0 and the elevation z, or else it will
be assumed to be constant (101.3 kPa). If no wind speed data U2 are available, a constant value of 2
m per second is used. Custom CO2 atmospheric concentration CO2 can also be provided, following
Yang et al. (2019).  

First all three PETs are calculated, but further we proceed with the PET_penman, because it is the most accurate one and we have all the data for it.

```{r evapotranspiration}

## Calculation of potential evapotranpiration  

sapply(monthly_climate, function(x) sum(is.na(x)))

monthly_climate <- monthly_climate %>%
  mutate(PET_penman = penman(Tmin = t_min, 
                             Tmax = t_max, 
                             U2 = wind_speed, 
                             CC = cloud_cover, 
                             lat = lat_station, 
                             z = elevation_station, 
                             P = air_pressure, 
                             RH = rel_humidity, 
                             ed = vapour_pressure,
                             # tsun = sunshine_hours # not complete data available!
                             na.rm=T),
         PET_hargreaves = hargreaves(Tmin = t_min, 
                                     Tmax = t_max, 
                                     lat = lat_station, 
                                     Pre = precipitation),
         PET_thornthwaite = thornthwaite(Tave = t_mean, 
                                         lat = lat_station))


# Then the Waterbalance is calculated as: Precipitation minus PET_penman:

monthly_climate <- monthly_climate %>% 
  mutate(D = precipitation-PET_penman)

# Converting the climate datatframe into a timeseries object, for the calculation of SPEI:
climate_ts <- ts(monthly_climate[,-c(1:2)], 
                 start=c(1973, 1),  
                 frequency=12) # means that I have monthly data input, so 12 month is one next time object (year)  


```


## 1.6. Calculation of SPEI

Calculation of the SPEI values in three different time scales: 12 months, 6 months and 3 months with reference period of 40 years from 1986 to 2025  

```{r calculation SPEI}
# caclcualte the SPEI for 3 different time scales with the spei() function
spei12 <- spei(climate_ts[,'D'], 
               scale=12, # time scale 12 means, the actual month plus the 11 before
               ref.start=c(1986,1), 
               ref.end=c(2025,12))

spei06 <- spei(climate_ts[,'D'], 
               scale=6, # time scale 6 means, the actual month plus the 5 before
               ref.start=c(1986,1), 
               ref.end=c(2025,12))

spei03 <- spei(climate_ts[,'D'], 
               scale=3, # time scale 3 means, the actual month plus the 2 before
               ref.start=c(1986,1), 
               ref.end=c(2025,12))

# extract the calculated values of SPEI
spei_12 <- as.numeric(spei12$fitted) 
spei_06 <- as.numeric(spei06$fitted)
spei_03 <- as.numeric(spei03$fitted)

# sum(is.na(spei_12)) 
# note: it is normal that the first 11 values are NA, because with scale=12 the spei function always takes the actual month plus the past 11 month into the calculation. (for scale=6, 5 NAs, for scale=3, 2 NAs respectively)


# add the spei variables to the original data frame
climate_spei <- 
  monthly_climate %>% 
  add_column(spei_12) %>% 
  add_column(spei_06) %>% 
  add_column(spei_03)

```


## 1.7. Supplem. Fig. S5

```{r SPEI Figures, fig.height = 3, fig.width = 5, fig.align = "center"}

# dataframe for SPEI-12 
spei_data_12 <- 
  climate_spei %>% 
  filter(year >= 1985) %>% 
  filter(month == 12) %>% # filter month 12, because there we have the month December, plus the 11 months before, i.e. the whole year (see spei calculation, scale)
  select(c(year, month, spei_12))

# dataframe for SPEI-6
spei_data_06 <-
  climate_spei %>% 
  filter(year >= 1985) %>% 
  filter(month == 9) %>% # filter month 9, because there we have the month September, plus the 5 months before, i.e. the growing season from April-September (see spei calculation, scale)
  select(c(year, month, spei_06))

# dataframe for SPEI-3
spei_data_03 <-
  climate_spei %>% 
  filter(year >= 1985) %>% 
  filter(month == 7) %>% # filter month 7, because there we have the month July, plus the 2 months before, i.e. the summer season from May-July (see spei calculation, scale)
  select(c(year, month, spei_03))



# make the figures:

fig_spei_12 <- spei_data_12 %>% 
  ggplot(aes(x = year, y = spei_12)) +
  
  geom_hline(yintercept = 0, color = "grey30") +
  geom_hline(yintercept = c(-1, 1), linetype = "dotted", color = "grey30") +
 
  geom_line(color = "grey", linewidth = 0.8) +
  
  geom_point(aes(fill = spei_12), shape = 21, size = 4, color = "black", show.legend = FALSE) +
  
  geom_hline(yintercept = 0, color = "black") +
  geom_hline(yintercept = c(-1, 1), linetype = "dotted", color = "black") +
  
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#4575b4", midpoint = 0) +
    
  ylim(-2, 2.3) +

  scale_y_continuous(name = "Drought Index (SPEI 12)", 
                     breaks = seq(-2.5, 2.5, 0.5), 
                     limits = c(-2.5, 2.5)) +
  
   scale_x_continuous(breaks = c(seq(from = 1985, to = 2025, by = 5)))+

  theme_classic2(base_size = 16) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_blank()
  )

fig_spei_12


fig_spei_06 <- spei_data_06 %>% 
  ggplot(aes(x = year, y = spei_06)) +
  
  geom_hline(yintercept = 0, color = "grey30") +
  geom_hline(yintercept = c(-1, 1), linetype = "dotted", color = "grey30") +

  geom_line(color = "grey", linewidth = 0.8) +
  
  geom_point(aes(fill = spei_06), shape = 21, size = 4, color = "black", show.legend = FALSE) +
  
  geom_hline(yintercept = 0, color = "black") +
  geom_hline(yintercept = c(-1, 1), linetype = "dotted", color = "black") +
  
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#4575b4", midpoint = 0) +
  
  ylim(-2, 2.3) +

  scale_y_continuous(name = "Drought Index (SPEI 6)", 
                     breaks = seq(-2.5, 2.5, 0.5), 
                     limits = c(-2.5, 2.5)) +
  
   scale_x_continuous(breaks = c(seq(from = 1985, to = 2025, by = 5)))+

  theme_classic2(base_size = 16) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_blank()
  )

fig_spei_06



fig_spei_03 <- spei_data_03 %>% 
  ggplot(aes(x = year, y = spei_03)) +
  
  geom_hline(yintercept = 0, color = "grey30") +
  geom_hline(yintercept = c(-1, 1), linetype = "dotted", color = "grey30") +

  geom_line(color = "grey", linewidth = 0.8) +
  
  geom_point(aes(fill = spei_03), shape = 21, size = 4, color = "black", show.legend = FALSE) +
  
  geom_hline(yintercept = 0, color = "black") +
  geom_hline(yintercept = c(-1, 1), linetype = "dotted", color = "black") +
  
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#4575b4", midpoint = 0) +
    
  ylim(-2, 2.3) +
  
  scale_y_continuous(name = "Drought Index (SPEI 3)", 
                     breaks = seq(-2.5, 2.5, 0.5), 
                     limits = c(-2.5, 2.5)) +
  
   scale_x_continuous(breaks = c(seq(from = 1985, to = 2025, by = 5)))+

  theme_classic2(base_size = 16) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_blank()
  )

fig_spei_03


Figure_S5 <- ggarrange(fig_spei_12, 
                       fig_spei_06, 
                       fig_spei_03, 
                       nrow=3, 
                       labels = c("(a)", "(b)", "(c)"))



```


# 2. Climate Station on site (ARBOfun)
## 2.1. Preparation of data frame

```{r load the data}

# load the weather data from the station on site ARBOfun - as 10min intervals

data_climate_2024 <- 
  read.csv(here("data", "raw", "02_data_climate_station_ARBOfun.csv")) %>% 
  # change the UTC time zone to the Europe/Berlin time zone
  mutate(
    date_time = ymd_hms(date_time, tz = "UTC") %>%
      with_tz("Europe/Berlin")
  )

```


## 2.2. VPDleaf from porometer

The porometer devices log with each measurement also the leaf-to-air VPD, which we need to relate it to ambient VPD.

```{r join stom con data}

# load the measurement data from the porometers to use the environemental variables
# that were measured by the device directly (VPDleaf)
stom_con <- read_csv(here("data", "processed", "data_cleaned.csv"))

# join the weather data to the main data
 data <- 
  stom_con %>% 
  rename(day = date) %>% 
  arrange(date_time) %>% 
  mutate(nearest_time = lubridate::round_date(date_time, unit = "10 mins")) %>% 
  left_join(data_climate_2024, 
            by = c("nearest_time" = "date_time")) %>%
  # create a date variable that is character
  mutate(day_chr= as.character(day)) %>% 
  # add new interpolated VPD column
  # For each leaf measurement time, look up the nearest two weather station measurements and linearly interpolate the air VPD, so     that each leaf measurement has a matching air VPD.
  mutate(
    VPD_air_interp = approx( # base R linear interpolation function with x and y and the new estimates for xout
      x = data_climate_2024$date_time, # the known time points from the weather station
      y = data_climate_2024$VPD, # these are the corresponding air VPD from the weather station
      xout = date_time # the time points of my leaf VPD measurements, for which we want extimated values
    )$y # since we only want the interpolated VPD values not x and y
  ) %>% 
  # species as a factor
  mutate(species=as.factor(species)) %>% 
  mutate(tree_id=as.factor(tree_id)) %>% 
    # add a variable that has 1, 2, 3 per tree individual
  group_by(species) %>%
  mutate(tree_id_num = as.factor(as.numeric(factor(tree_id)))) %>%
  ungroup() %>% 
  # create a unique leaf_id variable
  unite("leaf_id", tree_id, leaf_nr, sep="_", remove=F) %>% 
  group_by(day) %>%
  mutate(
    first_time = min(date_time, na.rm = TRUE),
    last_time  = max(date_time, na.rm = TRUE),
    first_measurement = format(min(date_time, na.rm = TRUE), "%H:%M:%S"),
    last_measurement  = format(max(date_time, na.rm = TRUE), "%H:%M:%S")
  ) %>%
  ungroup() %>% 
   group_by(day) %>% 
   mutate(trees_measured = n_distinct(tree_id))

```




## 2.3. Supplem. Fig. S1 - VPD measurement days

```{r VPD figure, fig.width=3, fig.height=3}

measurement_times <- data %>% 
  dplyr::select(date, first_time, last_time) %>% 
  distinct() %>%
  mutate(
    first_time = hms::as_hms(first_time),
    last_time  = hms::as_hms(last_time)
  )


for_vpd_fig <- data_climate_2024 %>%
  left_join(measurement_times, by = "date") %>% 
  mutate(
    time_hms = hms::as_hms(time),
    in_window = time_hms >= first_time & time_hms <= last_time
  )


fig_vpd <- 
  ggplot(for_vpd_fig,
       aes(x = time_hms, y = VPD, group = date)) +

  # full line in grey
  geom_line(color = "grey85", linewidth = 1) +

  # colored section within measurement window
  geom_line(data = subset(for_vpd_fig, in_window),
            aes(color = factor(date)),
            linewidth = 1) +

  scale_color_brewer(palette = "Set2") +

  labs(y = "Air VPD [kPa]", 
       x = "time (CET)",
       color = "Date") +

scale_x_time(
  expand = c(0, 0),
    breaks = scales::breaks_width("1 hour"),
  labels = scales::label_time("%H:%M")
) +

  theme_bw(base_size = 18) +

  theme(
    # axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
fig_vpd

rm(for_vpd_fig)

```

## 2.4. Supplem. Fig. S4 - VPD leaf vs. VPD air

```{r VPD leaf vs. VPD air, fig.height=10, fig.width=10}

# model of the VPDleaf vs VPDair
conversion_model <- lmer(
  VPD_air_interp ~ VPDleaf * species + (1|tree_id), 
  data = data)

summary(conversion_model)
anova(conversion_model)

# Estimated slopes per species
slopes_per_species <- emtrends(conversion_model, ~ species, var = "VPDleaf")

slopes_df <- as.data.frame(slopes_per_species)

#  Calculate R² per species
r2_per_species <- data %>%
  group_by(species) %>%
  group_modify(~ {
    model <- lmer(
      VPD_air_interp ~ VPDleaf + (1 | tree_id),
      data = .x
    )
    
    r2vals <- performance::r2(model)
    
    tibble(
      R2_marginal = r2vals$R2_marginal,
      R2_conditional = r2vals$R2_conditional
    )
  })

#  Merge R² back to leaf_data for plotting
data_for_figure <- data %>%
  left_join(r2_per_species, by = "species")


newdata <- data %>%
  group_by(species) %>%
  summarise(
    VPDleaf = list(seq(
      min(VPDleaf, na.rm = TRUE),
      max(VPDleaf, na.rm = TRUE),
      length.out = 200
    )),
    .groups = "drop"
  ) %>%
  tidyr::unnest(VPDleaf)

newdata$pred <- predict(
  conversion_model,
  newdata = newdata,
  re.form = NA
)
  
#  Figure:
conversion_fig <- 
  ggplot(data_for_figure, 
         aes(x = VPDleaf, 
             y = VPD_air_interp)) +
  geom_abline(slope=1, colour="lightgrey") +
  geom_point(aes(colour=as.factor(tree_individual)), 
             alpha = 0.4, ) +
  geom_line(data=newdata, 
            aes(x=VPDleaf, 
                y=pred), 
            color = "darkred") +
  geom_text(
    data = r2_per_species,
    aes(
      x = 6,    # fixed x position for all facets
      y = 4,    # fixed y position for all facets
      label = paste0("R² = ", round(R2_conditional, 2))
    ),
    hjust = 1, vjust = 1,
    inherit.aes = FALSE
  ) +
  scale_colour_manual(values= c("#4EA8DE", "#A29BFE", "#5B7065")) +
  facet_wrap(~species, ncol=5) +
  labs(
  x = expression("Leaf-to-air vapour pressure deficit (" * VPD[L] * "; kPa)"),
  y = expression("Atmospheric vapour pressure deficit (" * VPD[air] * "; kPa)")
  ) +
  theme_bw() +
    theme(legend.position = "none")

conversion_fig

ggsave(here("output", "Fig_S4_VPDleaf_vs_VPDair.png"), conversion_fig, height=10, width=8)

```



## 2.5. Supplem. Tab. S1 - Environmental Conditions

```{r supplementary table environmental conditions}

median_table <- 
  data %>% 
  group_by(day) %>% 
  filter(date_time >= first_time & date_time <= last_time) %>%
  ungroup() %>% 
  group_by(day, first_measurement, last_measurement, trees_measured) %>% 
  summarise(
        VPD = median(VPD, na.rm = TRUE),
        temperature = median(temperature, na.rm = TRUE),
        rel_humidity = median(rel_humidity, na.rm = TRUE),
        wind_speed = median(wind_speed, na.rm = TRUE),
        radiation_global = median(radiation_global, na.rm = TRUE),
        # fract_diff_radiation = median(diffuse_fraction, na.rm = TRUE),
        soil_moisture_15 = median(soil_watercontent_15, na.rm = TRUE),
        soil_moisture_50 = median(soil_watercontent_50, na.rm = TRUE),
    .groups = "drop"
  )

median_table

```

# 3. Soil Conditions

### 3.1. UFZ Drought Monitoring Germany

```{r load the data from UFZ}

# load the data from UFZ Dürremonitor

drought_intensity_upper <- terra::rast(here("data", "raw", "droughtintensity_SM_0_25cm_apr-oct.nc"))
drought_intensity_all <- terra::rast(here("data", "raw", "droughtintensity_SM_Lall_apr-oct.nc"))

drought_monthly_all <- terra::rast(here("data", "raw", "SMI_Gesamtboden_monatlich.nc"))
drought_monthly_upper <- terra::rast(here("data", "raw", "SMI_Oberboden_monatlich.nc"))


# coordinates of the site research arboretum ARBOfun
arbofun <- terra::vect(
  data.frame(
    lon = 12.500000,
    lat = 51.266667
  ),
  geom = c("lon", "lat"),
  crs = "EPSG:4326"
)



# extract the data needed

# helper function
extract_ufz <- function(raster, site, year = NULL) {
  
  if (!is.null(year)) {
    raster <- raster[[format(time(raster), "%Y") == year]]
  }
  
  raster |>
    terra::extract(site) |>
    as_tibble() |>
    pivot_longer(
      cols = -ID,
      names_to = "layer",
      values_to = "value"
    ) |>
    mutate(
      date = time(raster)
    )
}


## extract data needed

# Monthly SMI, 2024
smi_upper <- extract_ufz(
  drought_monthly_upper,
  arbofun,
  year = "2024"
)

smi_all <- extract_ufz(
  drought_monthly_all,
  arbofun,
  year = "2024"
)


# April–October drought intensity, 2024
drought_intensity_upper_site <- extract_ufz(
  drought_intensity_upper,
  arbofun,
  year = "2024"
)

drought_intensity_all_site <- extract_ufz(
  drought_intensity_all,
  arbofun,
  year = "2024"
)


terra::describe(drought_monthly_upper) |> head(50)

```




## 3.2. Supplem. Fig. S2

### 3.2.1. S2a - Drought Monitor 2024

```{r map}

smi_plot <- bind_rows(
  smi_upper  %>%  mutate(soil = "Upper soil (0–25 cm)"),
  smi_all  %>%  mutate(soil = "Total soil")
)

smi_fig <- ggplot(smi_plot, aes(x = date, y = value, linetype = soil)) +
  
  geom_hline(
    yintercept = c(0.20),
    linetype = "solid",
    linewidth = 0.7,
    colour="darkred"
  ) +
  geom_rect(
  aes(
    xmin = as.Date("2024-06-12"),
    xmax = as.Date("2024-07-09"),
    ymin = -Inf,
    ymax = Inf
  ),
  inherit.aes = FALSE,
  alpha = 0.02
  ) +
  
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  
    annotate(
      "text",
      x = as.Date("2024-01-01"),
      y = 0.20,
      label = "drought threshold",
      hjust = 0,
      vjust = -0.5,
      size = 3.5,
      colour="darkred"
    ) +
  
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2)
  ) +
  
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b"
  ) +
  
  labs(
    x = NULL,
    y = "Soil moisture index (SMI)",
    linetype = NULL
  ) +
  theme_classic(base_size=18)

```



### 3.2.2. S2b - Soil moisture daily

```{r soil moisture}

# prepare the dataframe
Soil_Moisture_2024 <- data_climate_2024 %>%
  dplyr::select(date_time, time, date, starts_with("soil")) %>% 
  pivot_longer(cols=starts_with("soil"),
               names_to = c(".value", "depth"),
               names_pattern = "(.+)_([0-9]+)$")

Soil_Moisture_Measurementdays <- 
Soil_Moisture_2024 %>% 
  filter(date=="2024-06-12" | date=="2024-06-13" | date=="2024-06-18" | date=="2024-06-24" | date=="2024-06-25" | date=="2024-06-26" | date=="2024-06-27" | date=="2024-07-09")

measurement_days <- Soil_Moisture_Measurementdays %>% distinct(date) %>% pull(date)
  



for_soil_fig <- data_climate_2024 %>%
    dplyr::select(date_time, time, date, starts_with("soil")) %>% 
  pivot_longer(cols=starts_with("soil"),
               names_to = c(".value", "depth"),
               names_pattern = "(.+)_([0-9]+)$") %>% 
    filter(date %in% measurement_days) %>%
  left_join(measurement_times) %>% 
  mutate(
    time_hms = hms::as_hms(time),
    in_window = time_hms >= first_time & time_hms <= last_time
  )

plot_data <- for_soil_fig %>% 
  filter(depth==50)



fig_soil_moist <- 
  plot_data %>% 
  ggplot(
       aes(x = time_hms, y = soil_watercontent, group = date)) +

  # full line in grey
  geom_line(color = "grey85", linewidth = 1) +

  # colored section within measurement window
# colored section within measurement window
  geom_line(data = subset(for_soil_fig, in_window & depth == 50), 
            aes(color = factor(date), group = date),             
            linewidth = 1) +

  scale_color_brewer(palette = "Set2") +

  labs(y = "Volumetric soil moisture content [%]", 
       x = "Time (CET)",
       color = "Date") +

scale_x_time(
  expand = c(0, 0),
    breaks = scales::breaks_width("1 hour"),
  labels = scales::label_time("%H:%M")
) +

  theme_classic(base_size = 18) +
ylim(13.2, 25.5) +
  theme(
    # axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
fig_soil_moist

```


### 3.2.3. S2 joined
```{r supplementary soil moisture joined, fig.width=9, fig.height=10}

# Stack Plot A over Plot B
Figure_S2 <- smi_fig /fig_soil_moist + 
  plot_layout(guides = "keep", design = "A\nB") + # Keeps a single legend if they share the same scale
  plot_annotation(
  tag_levels = "a",
  tag_prefix = "(",
  tag_suffix = ")"
)

Figure_S2

ggsave(here("output", "Fig_S2_Soil.png"), height=12, width=11)

```


## 3.3. Supplem Fig. S3 - Stratification of measurement days

```{r measurement days justification}

days_trees <- data %>% 
  dplyr::select(species, tree_id, day) %>% 
  distinct()

days_moist <- data %>% 
  dplyr::select(species, tree_id, day, soil_watercontent_50) %>% 
  group_by(species, tree_id, day) %>% 
  summarize(soil_moist = mean(soil_watercontent_50)) %>% 
  ungroup() %>% 
  group_by(species) %>% 
  mutate(mean_per_sp = mean(soil_moist)) %>% 
  ungroup()

mean_all_days <- data %>% 
  ungroup %>%  
  dplyr::select(soil_watercontent_50) %>% 
  summarize(mean_soil_moist = mean(soil_watercontent_50)) %>% 
  pull()


Figure_S3 <- days_moist %>% 
  ggplot(aes(species, soil_moist)) + 
  geom_hline(yintercept=mean_all_days, color="black", linewidth=0.5, linetype="dashed") + 
  geom_path(aes(colour=mean_per_sp), linewidth=1) + 
  geom_point(aes(colour=soil_moist), shape=19, size=2.5) +  
  ylab("Mean soil water content [%] on measurement day") + 
  xlab("Species") +
scale_colour_gradientn(
    name = "Soil water content [%]",
  colours = c(
    "#DCEAF7",  # very light blue
    "#6BAED6",  # medium blue
    "#08519C"   # dark blue
  )
) +
  theme_minimal() + 
  theme(
  axis.text.x = element_text(
    size = 7,
    angle = 90,
    vjust = 0.5,
    hjust = 1,
    face = "italic"
  ),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = "right"
  )
Figure_S3

ggsave(here("output", "Fig_S3_stratification_days.png"), Figure_S3, height=6, width=9)

```

